Viral-mediated Oct4 overexpression and inhibition of Notch signaling synergistically induce neurogenic competence in mammalian Müller glia

Version 1

Abstract

Abstract

Retinal Müller glia in cold-blooded vertebrates can reprogram into neurogenic progenitors to replace neurons lost to injury, but mammals lack this ability. While recent studies have shown that transgenic overexpression of neurogenic bHLH factors and glial-specific disruption of NFI family transcription factors and Notch signaling induce neurogenic competence in mammalian Müller glia, induction of neurogenesis in wildtype glia has thus far proven elusive. Here we report that viral-mediated overexpression of the pluripotency factor Oct4 (Pou5f1) induces transdifferentiation of wildtype mouse Müller glia into bipolar neurons, and synergistically stimulates glial-derived neurogenesis in parallel with Notch loss of function. Single cell multiomic analysis shows that Oct4 overexpression leads to widespread changes in gene expression and chromatin accessibility, inducing activity of both the neurogenic transcription factor Rfx4 and the Yamanaka factors Sox2 and Klf4. This study demonstrates that viral-mediated overexpression of Oct4 induces neurogenic competence in wildtype retinal Muller glia, identifying mechanisms that could be used in cell-based therapies for treating retinal dystrophies.

Introduction

The homeodomain factor Oct4 (Pou5f1) is a central component of transcriptional regulatory networks that maintain pluripotency in embryonic stem (ES) cells. Along with Sox2, Klf4, and Myc, Oct4 is one of the Yamanaka factors sufficient to induce the formation of induced pluripotent stem cells (iPSCs) from virtually all somatic cell types (Takahashi and Yamanaka, 2006). In contrast to these other factors, Oct4 is selectively expressed in ES cells and essentially absent from somatic cells, including adult stem cells from every tissue examined. The two clear exceptions are found in gonadal stem cells (Takashima et al., 2013; Uhlen et al., 2010) and during neural crest formation, where Oct4 expression is essential for the ability of these cells to generate diverse cell types (Morrison and Brickman, 2006; Zalc et al., 2021). This has raised the question of whether Oct4 overexpression in vivo, alone or in combination with other pluripotency factors, may confer some level of multipotency on somatic cells in the central nervous system.

Sustained overexpression of the Yamanaka factors Oct4, Sox2, Klf4, and Myc induces teratoma formation in many tissues (Abad et al., 2013; Ohnishi et al., 2014). Transient or low-level Yamanaka factor overexpression, in contrast, has been reported to induce cellular rejuvenation in both neurons and other cell types, although this occurs without clear induction of proliferation or conferring multipotency (Ocampo et al., 2016; C. Wang et al., 2021). However, several studies have directly investigated whether viral-mediated overexpression in CNS neural stem or glial cells of Oct4 alone, or in combination with subsets of other pluripotency factors, can induce multipotency and neurogenesis. Results from these studies are mixed, with Oct4 overexpression in neural stem cells reported to either not induce neurogenesis (Asadi et al., 2015; Dehghan et al., 2015; Sim et al., 2011), to promote formation of oligodendrocyte progenitors (OPCs) (Yu et al., 2021), or to promote remyelination in committed OPCs (Dehghan et al., 2016). Two studies reported that viral-mediated Oct4 overexpression, either alone (Niu et al., 2013) or in combination with other Yamanaka factors (Gao et al., 2016), could directly convert reactive astrocytes into neurons. However, these studies used the GFAP minipromoter to drive gene expression, which has been reported to show insert-dependent ectopic insert-dependent silencing in glia and activation in neurons (L.-L. Wang et al., 2021). This limitation makes these findings difficult to interpret, as additional methods of validating cell lineage were not used.

Two previous studies have reported injury-dependent induction of Oct4 in retinal Müller glial cells in zebrafish and mice. In zebrafish, where injury induces Müller glia to undergo conversion to neurogenic progenitors (Wan and Goldman, 2016), it was reported that Oct4 is essential for glial reprogramming to occur (Sharma et al., 2019). In mice, where neurogenic competence is rapidly and actively suppressed following injury, it was reported that Oct4 expression is transiently induced, but fully repressed within 24 hours following injury (Reyes-Aguirre and Lamas, 2016). This raises the possibility that Oct4 expression may be sufficient to confer neurogenic competence on retinal Müller glial cells. In this study, we assessed whether AAV-mediated overexpression of Oct4 could induce reprogramming of adult mouse Müller glia. While we did not detect any expression of Oct4 in either zebrafish or mouse Müller glia following injury, we use both genetic cell lineage and single-cell omics analyses to unambiguously demonstrate that viral-mediated overexpression of Oct4 selectively induces the formation of bipolar neurons from wildtype mouse Müller glia without inducing substantial levels of proliferation. We further show that Oct4 overexpression synergistically enhances existing levels of Müller glia-derived neurogenesis in the absence of Notch signaling. Single-cell multiomic analysis demonstrates that Oct4 cooperates with Sox2 and Klf4, which are constitutively expressed in Müller glia, to broadly alter the gene regulatory landscape and induce expression of genes such as the neurogenic transcription factor Rfx4. This demonstrates that substantial levels of in situ glia-to-neuron reprogramming in mammalian retina can be induced by viral overexpression of a key Yamanaka factor.

Results

Oct4 and Nanog are not detectably expressed in Müller glia or Müller glia-derived progenitor cells in either zebrafish or mice

To examine the expression of Oct4 and other pluripotency-associated genes in wildtype and neurogenic Müller glia, we analyzed bulk and single-cell RNA-Seq data previously generated by our group from zebrafish and mice (Hoang et al., 2020). Consistent with previous findings (Nelson et al., 2011; Surzenko et al., 2013), in mice we detected expression of Sox2 and Klf4 in resting and both NMDA-treated and light-damaged Müller glia, along with injury-induced expression of Myc. Likewise, in zebrafish we observe expression of sox2 and myca/b (but not klf4) in both resting and injured Müller glia (Fig. S1a,b). However, we did not detect expression of either Oct4 or its direct target Nanog in either species in any treatment condition (Loh et al., 2006). Bulk RNA-Seq data obtained from purified Müller glia detected very low levels of Oct4 (pou5f3), but not Nanog, expression in some injured zebrafish Muller glia, but neither Oct4 nor Nanog was detected in either resting or injured mouse Müller glia (Fig. S1a,b). Similarly, we did not observe any induction of Oct4 in snRNA-Seq data from mouse Müller glia made neurogenic by selective loss of function of Rbpj alone or in combination with Nfia/b/x (Le et al., 2024) (Fig. S1c). These results indicate that neither NMDA excitotoxicity, light damage, or induction of neurogenic competence by loss of function of Nfia/b/x and/or Rbpj induces detectable levels of Oct4 or Nanog in neurogenic mammalian Müller glia.

AAV-mediated overexpression of Oct4 induces conversion of wildtype mouse Müller glia to bipolar-like cells

To determine whether Oct4 overexpression could reprogram mammalian Müller glia into a neurogenic state, we administered intraperitoneal tamoxifen injections into P21 GlastCreER;Sun1-GFP mice, irreversibly labeling Müller glia with Sun1-GFP (de Melo et al., 2012; Hoang et al., 2020) (Fig. S2a, Fig. 1a). One week following the final tamoxifen dose, we performed intravitreal injections of either Gfap-Oct4-mCherry or Gfap-mCherry control 7m8 AAV. While expression of neurogenic bHLH factors driven by the Gfap promoter were rapidly silenced in Müller glia and activated in amacrine and retinal ganglion cells, Gfap-Oct4-mCherry remained selectively expressed in Müller glia but did not induce detectable levels of Müller glia-derived neurogenesis in the uninjured retina (Le et al., 2022). To assess whether excitotoxic injury could promote neurogenic competence, we injected N-methyl-d-aspartate (NMDA) (Hoang et al., 2020) one week post-viral injection. We then administered EdU following NMDA injection for one week to label proliferating cells. Retinas were harvested and analyzed at 1, 2, and 5 weeks post-injury (Fig. S2b, Fig. 1b).

AAV-mediated overexpression of Oct4 induces neurogenesis in wildtype mouse Müller glia following NMDA-induced excitotoxicity.

(A) Schematic of the transgenic construct used to specifically label Müller glia with Sun1-GFP expression. (B) Schematic of the GFAP AAV constructs and experimental workflow. (C) Representative images of retinas immunolabeled for GFP, mCherry, Otx2, and HuC/D. White arrowheads indicate GFP-positive Müller glia-derived neurons expressing neuronal markers Otx2 or HuC/D. (D) Quantification of mean percentage ± SD of GFP-positive Müller glia-derived neurons expressing either Otx2 or HuC/D. Significance was determined via two-way ANOVA with Tukey’s multiple comparison test: ****p < 0.0001. Each data point was calculated from an individual retina. TAM, tamoxifen; ONL, outer nuclear layer; INL, inner nuclear layer; GCL, ganglion cell layer. Scale bar = 50μm.

Immunostaining analysis revealed enriched mCherry expression in Müller glia in both control mCherry and Oct4-mCherry infected retinas at all timepoints (Fig. 1C, Fig. S2). In addition, Oct4 expression was detected in GFP-positive Müller glia in Oct4-mCherry infected retinas but not in mCherry control retina (Fig. S2i-j, S3i-j, S4a). We also observed selective expression of the neurogenic bHLH factor Ascl1 in Oct4-mCherry infected Müller glia at 1 week post-injury, with the number of Ascl-positive cells declining somewhat by 2 weeks post-injury (Fig. S2e,f). While Sun1-GFP-positive cells infected with Oct4-mCherry mostly express the Müller glial marker Sox2 (Fig. S2a,b), from 2 weeks post-injury onwards a subset of GFP-positive cells did not show detectable Sox2 expression (Fig. S2b, yellow arrows).

At 1 week post-injury, we did not observe any significant Müller glia-derived neurogenesis. By 2 weeks post-injury, 1.1% of GFP-positive cells co-localized with the bipolar cell marker Otx2 in Oct4-mCherry infected retinas (Fig. S2k,l,n), and this increased to 5.1% of GFP-positive cells at 5 weeks post-injury (Fig. 1c,d). We observed that Müller glia-derived Otx2+ neurons show little or no mCherry reporter expression, due to reduced activity of the Gfap minipromoter in neurons. Immunostaining with additional cell-specific markers revealed that a subset of these GFP-positive cells expressed the cone bipolar markers Scgn and Cabp5 (Fig. S3f,g), but not the rod and cone ON bipolar cell marker Isl1 or the rod marker Nrl (Fig. S3d,e). Furthermore, we did not observe expression of the amacrine cell marker HuC/D in GFP-positive cells (Fig. 1c,d), which is observed in neurons generated from neurogenic mouse Müller glia deficient in either Rbpj or Nfia/b/x (Hoang et al., 2020; Le et al., 2024).

To test whether Oct4 overexpression could induce proliferation in wildtype Müller glia, we administered 5-ethynyl-2’-deoxyuridine (EdU) through daily intraperitoneal injections and drinking water following NMDA injury. At 2 weeks post-injury, we observed a very limited number of GFP-positive cells incorporating EdU in retinas infected with both mCherry control and Oct4-mCherry, with EdU incorporation in the second post-injury week modestly but significantly increased by Oct4 overexpression (Fig. S2f,n). However, no EdU/GFP-positive cells were detected at either 1 week (Fig. S2e,n) or 5 weeks post-injury (Fig. S4h). These findings indicate that Oct4 overexpression induces a very low level of proliferation in wildtype Müller glia in the second week following NMDA injury. However, the absence of EdU/Otx2/GFP-positive cells at both 2 and 5 weeks post-injury suggests that bipolar cell generation at this stage was, at least in part, driven by direct transdifferentiation.

Oct4 overexpression synergistically enhances glial-derived neurogenesis induced by genetic disruption of Notch signaling

In our previous study, we demonstrated that disrupting Notch signaling in adult mouse Müller glia–either by selective deletion of the transcriptional mediator Rbpj or by combined disruption of Notch1 and Notch2 receptors–induces transdifferentiation into bipolar and amacrine-like cells (Le et al. 2024). To further determine the effects of Oct4 overexpression in the absence of Notch signaling, we used GlastCreER;Rbpjlox/lox;Sun1-GFP mice to genetically label Müller glia while disrupting Rbpj function (Fig. 2a, S4a). We then performed intravitreal injection of either control or Oct4 overexpression constructs, followed by NMDA and EdU injection as performed for GlastCreER;Sun1-GFP mice. Retinas were collected at 1, 2 and 5 weeks following NMDA treatment (Fig. 2b, S4b). Immunostaining confirmed expression of Oct4 protein in GFP-positive Müller glia of Oct4-mCherry infected retinas, but not in mCherry-infected controls (Fig. S3j, S4i,j). We observed that many GFP-positive cells in Oct4-mCherry-infected retinas did not express the Müller glial markers Sox2 (Fig.S4c,d) and Sox9 (Fig. S3j, yellow arrows), suggesting that they have lost their glial identity.

Oct4 overexpression enhances neurogenesis observed in Rbpj-deficient Müller glia.

(A) Schematic of the transgenic constructs used to induce loss of function of Rbpj specifically in Müller glia. (B) Schematic of the GFAP AAV constructs and experimental workflow. (C) Representative images of retinas immunolabeled for GFP, mCherry, Otx2, and HuC/D. White arrowheads indicate GFP-positive Müller glia-derived neurons expressing neuronal markers Otx2 or HuC/D. (D) Quantification of mean percentage ± SD of GFP-positive Müller glia-derived neurons expressing either Otx2 or HuC/D. Significance was determined via two-way ANOVA with Tukey’s multiple comparison test: ****p < 0.0001. TAM, tamoxifen; ONL, outer nuclear layer; INL, inner nuclear layer; GCL, ganglion cell layer. Scale bar = 50μm.

Five weeks following NMDA injury, we observed 4.5% of GFP-positive cells in retinas of mCherry-infected controls expressed the bipolar cell marker Otx2, while 0.7% expressed the amacrine cell marker HuC/D (Fig. 2c,d). These levels of neurogenesis are consistent with previous findings in uninfected Rbpj-deficient Müller glia (Le et al., 2024). In contrast, we observe 24.3% of GFP-positive cells expressing Otx2 in Oct4-mCherry-infected retinas, representing a significant (p < 0.0001) increase relative to control. However, we did not observe a significant change in the relative fraction of GFP/HuC/D-double-positive cells. In contrast to wildtype retinas, no EdU incorporation was detected in Oct4-expressing, Rbpj-deficient Müller glia at either 1 or 2 weeks following infection (Fig. S4e,f), indicating that the observed neurogenesis is due largely to direct transdifferentiation of Müller glia into retinal neuron-like cells, rather than through dedifferentiation to a proliferative neuronal progenitor-like state.

Next, we tested whether Oct4 overexpression could further enhance neurogenesis in other cell-specific mutants that induce neurogenic competence in adult Müller glia. We have previously observed Müller glia-specific Notch1/2 double mutants phenocopy neurogenesis seen in Rbpj mutants (Le et al., 2024). To do this, we used GlastCreER;Notch1lox/lox;Notch2lox/lox;Sun1-GFP mice (Fig. S5a), and analyzed the effect of Oct4 overexpression using the same protocol previously described for wildtype and Rbpj mutants (Fig. S5b). We observed 15.8% of GFP+ Müller glia colocalized with Otx2 in Oct4-mCherry-infected retinas compared to 6.6% GFP/Otx2+ cells in samples infected with control mCherry vector (Fig. S5c). Infection with Oct4-mCherry vector induced both Oct4 (Fig. S5e) and Ascl1 (Fig. S5d) expression in Notch1/2-deficient Müller glia.

Finally, we also tested whether Oct4 overexpression could enhance Müller glia-derived neurogenesis induced by loss of function of the transcription factors Nfia, Nfib, and Nfix (Hoang et al., 2020). GlastCreER;Nfialox/lox;Nfiblox/lox;Nfixlox/lox;Sun1-GFP mice (Fig. S6a) were infected with both mCherry and Oct4-mCherry vectors as previously described, and harvested at 5 weeks following NMDA infection (Fig. S6b). However, we observed only very limited coexpression of mCherry and GFP following infection with either construct (Fig. S6c), indicating that NFI factors may be required for appropriate glial-specific expression of the GFAP minipromoter construct, as suggested by previous studies in astrocytes (Cebolla and Vallejo, 2006).

Oct4 overexpression promotes expression of reactive glial markers and neurogenic bHLH factors

To gain insight into the mechanism by which Oct4 stimulates neurogenesis in Rbpj-deficient Müller glia, we conducted single-cell (sc)RNA-Seq analysis of GFP-positive cells FACS-isolated from retinas of GlastCreER;Rbpjlox/lox;Sun1-GFP mice infected with either GFAP-mCherry control or GFAP-Oct4-mCherry virus. Retinas were tamoxifen- and NMDA-treated as described in Figure 2, and GFP-positive cells isolated at 8 weeks following NMDA treatment (Fig. 3a). We profiled 12,143 GFP+ cells from control-infected and 14,612 GFP+ cells from Oct4-mCherry-infected retinas (Fig. 3b). In both samples, we observe a clear differentiation trajectory connecting Müller glia, neurogenic Müller glia-derived progenitor cells (MGPCs), and differentiating amacrine and bipolar cells (Fig. 3b). Oct4-infected retinas consistently show substantially higher fractions of bipolar cells (20.5 % vs. 2.7 %), and a smaller fraction of Müller glia (71.4 % vs. 88.2 %), consistent with immunohistochemical data (Fig. 3c).

ScRNA-Seq analysis of Müller glia and Müller glia-derived neurons derived from Rbpj-deficient Müller glia following Oct4 overexpression.

(A) Schematic of the scRNA-seq experimental pipeline. (B) UMAP plot showing the clustering of GFP-positive cells from retinas infected with GFAP-mCherry and GFAP-Oct4-mCherry AAV constructs. (C) Stacked bar plots showing the proportion of cells in each cluster across two sample groups. (D) Feature plots highlighting the cluster of Müller glia (Hes5), neurogenic MGPCs (Neurog2, Insm1, Ascl1), bipolar cells (Otx2), and amacrine cells (Elavl3). (E) Heatmap showing the expression of top differentially expressed genes (DEGs) for Rbpj-deficient Müller glia cell cluster from retinas infected with GFAP-mCherry and GFAP-Oct4-mCherry AAV constructs. (F) Heatmap showing the expression of top differentially expressed genes (DEGs) for MGPC cell cluster from retinas infected with GFAP-mCherry and GFAP-Oct4-mCherry AAV constructs.

Selective molecular markers could readily distinguish Müller glia (Glul, Apoe, Vim, Clu, Rlbp1) from neurogenic MGPCs (Neurog2, Ascl1, Sox4, Hes6, Dll1), and clearly identified differentiating bipolar (Otx2, Pcp2, Pcp4, Car10, Gabrb3) and amacrine-like cells (Nrxn1, Tfap2b, Nrg1, Snap25) (Fig. 3d-f, Fig.S7, Table ST1). In both MGs and MGPCs, we observed substantial differences in gene expression between samples infected with control and Oct4-mCherry samples. Oct4-infected Müller glia selectively downregulated genes specific to resting Müller glia (Car2, Dkk3, Glul, Col23a1, Abca8a), and upregulated genes enriched in reactive glia (Gfap, Socs3, Nupr1, Mt1/2/3, Fos, Jun, Klf6) (Fig. 3e,f, Table ST1). In MGPCs, Oct4 also downregulated both transcription factors (Lhx2, Hopx) and other genes (Apoe, Dio1, Il33, Espn) that are enriched in resting Müller glia, and upregulated many genes selectively expressed in activated glia (Mt1/2/3, Fos, Jun, Nupr1, Socs3). In addition, Oct4-infected MGPCs expressed consistently higher levels of neurogenic bHLH factors (Ascl1, Neurog2, Hes6), bipolar cell specification (Otx2), or otherwise broadly neurogenesis (Insm1, Dll3, Mybl1) or are selectively expressed in late-stage neurogenic progenitors in the developing retina (Gadd45a, Btg2) (Clark et al., 2019) (Fig. 3e,f, Table ST1). These findings imply that Oct4 enhances bipolar cell formation in Rbpj-deficient Müller glia potentially by repressing expression of genes that promote quiescence such as Lhx2 (de Melo et al., 2018, 2016), increasing expression of genes that broadly promote neurogenesis, and selectively increasing expression of the bipolar cell-promoting factor Otx2 (Chan et al., 2020; Wang et al., 2014).

Single-cell multiomic analysis identifies targets of Oct4 in wildtype and Rbpj-deficient Müller glia

To further investigate the mechanism by which Oct4 promotes neurogenesis, we conducted simultaneous single-nucleus (sn)RNA-Seq and ATAC-Seq on GFP-positive cells from both wildtype and Rbpj-deficient Müller glia infected with either control or Oct4-mCherry vectors at 8 weeks following NMDA injection (Fig. 4a). UMAP analysis of integrated cellular snRNA/ATAC-Seq profiles was then used to identify cell clusters. Müller glia from wildtype and Rbpj-deficient animals form two separate Müller glia clusters (Fig. 4b,c). AAV-infected wildtype Müller glia could be readily identified by selective expression of Oct4 (Fig.4e). Wildtype Oct4-expressing Müller glia give rise to both small numbers of neurogenic MGPCs (Fig. 4b), in line with the small numbers of Ascl1-positive cells observed (Fig. S2g,h), as well as mature bipolar cells, consistent with histological findings (Fig. 1c,d). Oct4 overexpression substantially enhances bipolar cell generation from Rbpj-deficient MGPC (Fig. 4d, Fig. S8). Although small numbers of rod and cone photoreceptors are detected in this analysis, their proportions do not substantially change between treatment conditions, and therefore likely represent cellular contaminants.

Integrated snRNA/scATAC-Seq analysis of control and Rbpj-deficient Müller glia with Oct4 overexpression.

(A) Schematic of the multiomic scRNA/ATAC-seq experimental pipeline. (B) UMAP plot of multiomic datasets showing the clustering of GFP+ cells from control and Rbpj-deficient Müller glia from GFAP-mCherry and GFAP-Oct4-mCherry AAV infected retinas. (C) UMAP plot showing the identity of cell clusters determined by marker gene expression. (D) Stacked bar plots represent the proportion of cells in each cluster across different sample groups. (E) Feature plots highlighting the cluster of Müller glia (Hes1), neurogenic MGPCs (Neurog2, Sox4), bipolar cells (Otx2), Pou5f1- and Rfx4-expressing cells. (F) Heatmap showing expression of DEGs along the neurogenesis trajectory. (G) Heatmap showing differential motif activity along the neurogenesis trajectory. (H) Heatmaps of DEGs differentially expressed between Oct4 and mCherry control samples in the wildtype and Rbpj-deficient Müller glia cell clusters. Venn diagram showing unique and common DEGs between Oct4 and mCherry control in the wildtype and Rbpj-deficient Müller glia clusters. (I) Increased chromatin accessibility regions associated with the Oct4 and Rfx4 loci observed following Oct4 overexpression in both wildtype and Rbpj-deficient Müller glia. (J) Top-ranked enriched motifs in chromatin regions showing increased accessibility following Oct4 overexpression in both wildtype and Rbpj-deficient Müller glia cell clusters.

Oct4 overexpression leads to significant changes in gene expression in both wildtype and Rbpj-deficient Müller glia. The transition from resting to neurogenic MGPCs to bipolar cells is associated with downregulation of genes that are specifically expressed in resting Müller glia, including Aqp4, Hes1, Tcf7l2, and Lhx2 (Fig. 4e,f, Table ST2). Similarly, motif analysis showed that chromatin accessibility is reduced for target motifs for transcription factors that maintain glial quiescence such as Nfia/b/x and Lhx2 (Fig. 4g, Fig. S8, Table ST2). MGPCs likewise show increased accessibility at Stat1/3 motifs, as well as target motifs for neurogenic transcription factors such as Ascl1, Neurog2, and Neurod1 (Fig. 4g), consistent with previous studies of MGPCs generated from Nfia/b/x and Rbpj-deficient Müller glia (Hoang et al., 2020; Le et al., 2024).

Oct4 overexpression in both wildtype and Rbpj-deficient Müller glia clusters respectively induces upregulation of 534 and 1657 genes, while respectively downregulating 439 and 824 genes (Fig. 4h). A total of 156 genes are upregulated in both wildtype and Rbpj-deficient Müller glia, while 63 genes are downregulated. In both wildtype and Rbpj-deficient Müller glia, Oct4 upregulates multiple classes of ion channels, including mechanosensitive (Piezo2), TRP (Trpm1, Trpm3), and chloride channels (Clca3a1, Clca3a2), as well as GABA receptors (Gabra4, Gabrb1), cyclic nucleotide phosphodiesterases (Pde8b, Pde11a), and the neuropeptide Npy. Oct4 overexpression also selectively downregulates multiple inflammation-associated genes (B2m, H2-D1), small GTPases (Gbp2, Igtp, Irgm2, Rab13, Rgs20, Tgtp2), along with the glycolytic enzyme Gapdh and many ribosomal subunits (Fig. 4h, Table ST3). This functionally diverse range of genes that are not prominently expressed in retinal progenitors, and whose role in controlling neurogenic competence remains to be investigated.

Oct4 also differentially regulates multiple transcription factors in both wildtype and Rbpj-deficient Müller glia. As expected, Pou5f1 mRNA is strongly upregulated (Fig. 4e,h), with increased chromatin accessibility also being observed at the endogenous Pou5f1 promoter, indicating that Oct4 regulates its self-expression (Fig. 4i, Table ST4). Similarly, Oct4 overexpression strongly upregulates the transcription factor Rfx4 and accessibility at its endogenous promoter (Fig. 4e,h,i, Table ST4). Overexpression of Rfx4 is sufficient to induce neurogenesis when overexpressed in human embryonic stem cells (Choi et al., 2024; Joung et al., 2024). However, Rfx4 is not detectably expressed in developing mouse retinas (Clark et al., 2019), indicating that Oct4-induced neurogenesis may involve mechanisms beyond simply inducing dedifferentiation to a retinal progenitor-like state.

Several other transcription factors are only differentially regulated following Oct4 overexpression in Rbpj-deficient Müller glia. One notable upregulated transcription factor is Klf4, which cooperates with Oct4 to induce pluripotency (Takahashi and Yamanaka, 2006). In contrast, Stat1, which inhibits neurogenesis induced by Ascl1 overexpression in Müller glia (Jorstad et al., 2020), is selectively downregulated in Rbpj-deficient Müller glia (Fig. 4h, Table ST3).

Analysis of changes in chromatin accessibility induced by Oct4 reveals that the consensus motif for Oct4 and its close variant Pit1 (Pou1f1) show the strongest increase in accessibility, followed by Rfx family motifs (Fig. 4j, Table ST5). Both wildtype and Rbpj-deficient Müller glia show increased accessibility associated with Sox2, which directly interacts with Oct4 and cooperates to induce pluripotency (Rizzino, 2013). While Oct4 does not significantly upregulate the pluripotency factor Klf4 in wildtype Müller glia, it robustly induces Klf4 expression Rbpj-deficient Müller glia, and likewise selectively induces accessibility at Klf4 consensus motif sites (Fig. 4j). In addition, Rbpj-deficient Müller glia also show smaller but still significant increases in accessibility at target sites for the proneural factor Neurog2 and the neurogenic factor Insm1 (Lyu et al., 2021), identifying additional mechanisms by which Oct4 might synergistically enhance neurogenesis (Fig. 4j, Table ST5).

Discussion

In this study, we demonstrate that AAV-mediated overexpression of Oct4 is sufficient to induce neurogenic competence in wildtype Müller glia in adult mice, and to synergistically enhance neurogenesis induced by genetic disruption of Notch signaling. While Oct4 overexpression induces very low levels of proliferation, Oct4-induced neurogenesis appears to mostly result from direct glia-to-bipolar cell transdifferentiation by simultaneously upregulating expression of neurogenic bHLH factors and the bipolar cell promoting factor Otx2. In addition, Oct4 upregulates expression of the neurogenic transcription factor Rfx4, which is not expressed during retinal development. Oct4 exerts reprogramming effects by broadly increasing chromatin accessibility at predicted regulatory sites associated with these genes.

While we did not detect endogenous Oct4 expression in either injured or neurogenic mouse Müller glia, the fact that these cells express robust levels of Sox2, Klf4, and Myc implies that they may already be prime to reprogramming induced by Oct4. Our findings differ from two previous reports that initially motivated us to carry out this study. A previous study in zebrafish showed weak levels of the Oct4 homolog pou5f3 in resting glia, with these levels increasing dramatically following injury (Sharma et al., 2019). It remains unclear why we were unable to detect more than trace levels of pou5f3 expression in zebrafish Müller glia in any of the many bulk and single-cell RNA-Seq samples we analyzed (Hoang et al., 2020). We also did not detect the previously reported low levels of injury-induced Oct4 expression in mouse Müller glia (Reyes-Aguirre and Lamas, 2016). Similar discrepancies in observed cellular patterns of Oct4 expression have been observed in many other tissues (Lengner et al., 2008), and further work will be needed to clarify these discrepancies.

This study represents a clear instance in which viral-mediated expression coupled with genetic cell lineage analysis was effectively used to induce the conversion of adult mammalian Müller glia into retinal neurons in situ. A similar approach has been also successfully used to generate bipolar neurons from both immature (Pollak et al., 2013) and very recently also in mature, Müller glia by overexpression of the neurogenic bHLH factors Ascl1 and Atoh1 (Pavlou et al., 2024). Other studies reported efficient conversion of Müller glia to photoreceptors using a complex cocktail of overexpressed factors, including Ctnnb1, Otx2, Crx, and Nrl (Yao et al., 2018), and efficient conversion of Müller glia to retinal ganglion cells following overexpression of Atoh7 and Pou4f2 (Xiao et al., 2021). However, these studies both used GFAP minipromoters to both drive expression of these constructs and to infer cell lineage. Multiple studies have shown that GFAP minipromoter specificity is heavily influenced by insert sequences present in AAVs, which can result in both silencing of expression in glia and ectopic activation of expression in neurons (Le et al., 2022; L.-L. Wang et al., 2021). As a result, it cannot be concluded that these represent bona fide glia-to-neuron conversion in vivo. Similar claims have been made about knockdown of the RNA binding protein Ptbp1, which was variously claimed to directly convert Müller glia to either retinal ganglion cells or photoreceptors, but has been debunked using genetic cell lineage analysis (Hoang et al., 2022; Xie et al., 2022). In this study, the combination of prospective cell lineage analysis using the GlastCreER;Sun1-GFP transgene before infection with GFAP-Oct4-mCherry vector, in combination with scRNA-Seq analysis of infected cells, rules out any potentially confounding factors resulting from the GFAP minipromoter.

Our finding that Oct4 can synergistically enhance neurogenesis induced by genetic disruption of Notch signaling raises the question of whether this might be more broadly applicable to other models of induced neurogenesis in mammalian Müller glia. The mechanism by which Oct4 induces glial-derived neurogenesis appears complex, and to differ fundamentally from that seen following either overexpression of Ascl1 or loss of function of Nfia/b/x and/or Rbpj (Hoang et al., 2020; Jorstad et al., 2017; Le et al., 2024). In these cases, Müller glia broadly upregulate proneural genes and/or downregulate Notch signaling. Oct4 instead induces expression of the neurogenic transcription factor Rfx4, which is not expressed in developing retina. It is likely that activation of this parallel pathway to neurogenic competence in part accounts for synergistic induction of neurogenesis seen in Rbpj-deficient Müller glia.

Other mechanisms enhancing Oct4-dependent neurogenesis in the Rbpj-deficient Müller glia may include enhanced induction of Klf4 expression and increased accessibility at target sites for neurogenic transcription factors such as Ascl1. While Ascl1 overexpression induces Müller glia-derived neurogenesis in adult mice, this occurs only following treatment with HDAC inhibitors (Jorstad et al., 2017), even though Ascl1 is itself a pioneer factor (Păun et al., 2023). Oct4 overexpression broadly increases the accessibility of putative cis-regulatory sequences associated with both Neurog2 and Insm1, identifying another potential mechanism of synergy.

Transient overexpression of Oct4 could potentially further enhance the neurogenic function of Ascl1 and enhance glial-derived neurogenesis more broadly. This may also be the case for pathways reported to induce glial reprogramming in other CNS regions, such as Sox2 overexpression (Niu et al., 2015). Since Oct4 overexpression only weakly and transiently induces proliferation in adult Müller glia, this reduces the potential oncogenic risk, and may ultimately prove to be relevant for the design of cell-based regenerative therapies for treating retinal dystrophies.

Methods

Mice

Mice were raised and housed in a climate-controlled pathogen-free facility on a 14/10 h light/dark cycle. Mice used in this study were GlastCreER;Sun1-GFP, which were generated by crossing the GlastCreER and Sun1-GFP lines developed by Dr. Jeremy Nathans at Johns Hopkins (de Melo et al., 2012; Mo et al., 2015), and were obtained from his group.

GlastCreER;Rbpjlox/lox;Sun1-GFP mice were generated by crossing GlastCreER;Sun1GFP with conditional Rbpjlox/lox mice (JAX #034200). GlastCreER;Notch1lox/lox;Notch2lox/lox;Sun1-GFP mice were obtained by crossing GlastCreER;Notch1lox/lox;Sun1-GFP and GlastCreER;Notch2lox/lox;Sun1-GFP mice. GlastCreER;Nfia/b/xlox/lox;Sun1-GFP mice were generated by crossing Nfialox/lox(see below); Nfiblox/lox (Hsu et al., 2011), and Nfixlox/lox(Campbell et al., 2008) mice to GlastCreER;Sun1-GFP mice. Nfialox/lox mice were generated in the Roswell Park Gene Targeting and Transgenic Shared Resource using heterozygous targeted ES cells from EUCOMM project 38437 (KOMP).

Maintenance and experimental procedures performed on mice were in accordance with the protocol approved by the Institutional Animal Care and Use Committee (IACUC) at the Johns Hopkins School of Medicine.

Intraperitoneal tamoxifen (TAM) Injection

To induce Cre recombination, animals at ∼3 weeks of age were intraperitoneally injected with tamoxifen (Sigma-Aldrich, #H6278-50mg) in corn oil (Sigma-Aldrich, #C8267-500ML) at 1.5mg/dose for five consecutive days.

NMDA treatment

Adult mice were anesthetized with isoflurane inhalation. Two microliters of 100 mM NMDA in PBS were intravitreally injected using a microsyringe with a 33G blunt-ended needle.

EdU treatment

At 1-week post tamoxifen induction, mice were administered EdU (Abcam, #ab146186) via i.p. injections (150μl of 5mg/ml in PBS) and drinking water (0.3 mg/ml) for 7 consecutive days.

Cloning, production, and intravitreal injection of Adeno-Associated Virus

The Addgene #50473 construct which contains a GFAP promoter was used in this study. The EGFP sequence was replaced by the mCherry sequence. The P2A ribosomal self-cleaving peptide is used to simultaneously express Oct4 5′ to the mCherry reporter as a single polypeptide, which is then cleaved to generate Oct4 and mCherry. The coding sequence of Oct4 was synthesized by GeneWiz. AAV constructs were packaged into AAV2.7m8 by Boston Children’s Hospital Viral Core. Following tam induction, one-month-old GlastCreER;Sun1-GFP mice were intravitreally injected with GFAP AAV constructs using a microsyringe with a 33G blunt-ended needle. The titres and injection volume for each construct are listed in table ST6.

Immunohistochemistry and imaging

Collection and immunohistochemical analysis of retinas were performed as described previously (Hoang et al., 2020). Mouse eye globes were fixed in 4% paraformaldehyde (ElectronMicroscopySciences, #15710) for 4 h at 4°C. Retinas were dissected in 1x PBS and incubated in 30% sucrose overnight at 4°C. Retinas were then embedded in OCT (VWR, #95057-838), cryosectioned at 16 μm thickness, and stored at −20°C. Sections were dried for 30 min in a 37°C incubator and washed 3 × 5 min with 0.1% TritonX-100 in PBS (PBST). EdU labeling was performed by using Click-iT EdU kit (ThermoFisher, #C10340, #C10636) following the manufacturer’s instructions. Sections were then incubated in blocking buffer (10% Horse Serum (ThermoFisher, #26050070), 0.4% TritonX-100 in 1x PBS) for 2 h at room temperature (RT) and then incubated with primary antibodies in the blocking buffer overnight at 4°C. Primary antibodies used are listed in Table ST7.

Sections were washed 4 x 5 min with PBST to remove excess primary antibodies and were incubated in secondary antibodies in blocking buffer for 2 h at RT. Sections were then counterstained with DAPI in PBST, washed 4 x 5 min in PBST and mounted with ProLong Gold Antifade Mountant (Invitrogen, #P36935) under coverslips (VWR, #,48404-453), air-dried, and stored at 4°C. Fluorescent images were captured using a Zeiss LSM 700 confocal microscope. Secondary antibodies used are listed in table ST7.

Cell quantification and statistical analysis

Otx2/GFP-positive and HuC/D/GFP-positive cells were counted and divided by the total number of GFP-positive cells from a single random whole section per retina. Each data point in the bar graphs was calculated from an individual retina. All cell quantification data were graphed and analyzed using GraphPad Prism 10. Two-way ANOVA was used for analysis between 3 or more samples of multiple groups. All results are presented as mean ± SD.

Retinal cell dissociation

Retinas were dissected in fresh ice-cold PBS and retinal cells were dissociated using an optimized protocol as previously described (Fadl et al., 2020). Each sample contains a minimum of 4 retinas from 4 animals of both sexes. Dissociated cells were resuspended in ice-cold HBAG Buffer containing Hibernate A (BrainBits, #HALF500), B-27 supplement (ThermoFisher, #17504044), and Glutamax (ThermoFisher, #35050061).

Single-cell RNA-sequencing (scRNA-seq) library preparation

scRNA-seq were prepared on dissociated retinal cells using the 10X Genomics Chromium Single Cell 3’ Reagents Kit v3.1 (10X Genomics, Pleasanton, CA). Libraries were constructed following the manufacturer’s instructions and were sequenced using Illumina NextSeq. Sequencing data were processed through the Cell Ranger 7.0.1 pipeline (10X Genomics) using default parameters.

Single-cell Multiome ATAC + GEX sequencing library preparation

scATAC-seq and scRNA-seq were prepared on FACS-isolated GFP-positive cells using the 10X Genomic Chromium Next GEM Single Cell Multiome ATAC + Gene Expression kit following the manufacturer’s instructions. Briefly, cells were spun down at 500xg for 5 min, resuspended in 100 μl of ice-cold 0.1x Lysis Buffer, lysed by pipette-mixing 4 times, and incubated on ice for 4 min total. Cells were washed with 0.5 ml of ice-cold Wash Buffer and spun down at 500xg for 5 min at 4°C. Nuclei pellets were resuspended in 10-15 μl Nuclei Buffer and counted using Trypan blue. Resuspended cell nuclei (10-15k) were utilized for transposition and loaded into the 10x Genomics Chromium Single Cell system. ATAC libraries were amplified with 10 PCR cycles and were sequenced on Illumina NovaSeq with ∼200 million reads per library. RNA libraries were amplified from cDNA with 14 PCR cycles and were sequenced on Illumina NovaSeq 6000

scRNA-seq data analysis

ScRNA-Seq data were pre-processed using Cellranger v7.1.0 using custom mouse genome mm10 with mCherry and Sun1GFP sequences. The cell-by-genes count matrices were further analyzed using Seurat V5.1.0 (Hao et al., 2024).Cells with RNA counts less than 1000 or greater than 25,000, and number of genes less than 500 or greater than 6000 were filtered out as low-quality cells. Additionally, cells with a mitochondrial fraction of greater than 10% were removed. For data visualization, UMAP was generated using the first 20 dimensions. Müller glia and MGPC cell clusters were subsetted for further analysis. Differential gene expression of mCherry control and Oct4-mCherry groups was performed using “FindAllMarkers” function.

Single-cell Multiomic analysis

For the data processing, raw scRNA-seq and scATAC-seq data were processed with the Cell Ranger software version 2.0.2 specifically using cellranger-arc pipeline for formatting reads, demultiplexing samples, genomic alignment, and generating the cell-by-gene count matrix and the fragments files. Both the cell-by-gene count matrix and the fragments files are the final output from the cellranger-arc pipeline and were used for all downstream analysis.

For data analysis, scRNA-seq and scATAC-seq were analyzed using Archr version 1.0.2. EnsDb.Mmusculus.v79 and BSgenome.Mmusculus.UCSC.mm10 were used for annotations. The arrow file for each sample was created using the cell-by-gene count matrix and the fragments files output from the cellranger-arc pipeline. To filter poor quality cells, Gex_nGenes greater than 500 and less than 5000 were kept, Gex_nUMI greater than 1000 and less than 15,000 were kept. Also, TSSEnrichment greater than 10 and nFrags greater than 5000 were kept, other cells were deemed of poor quality and were filtered out. Gene expression values from a paired scATAC-seq and scRNA-seq multi modal assay were added to the arrow files using “addGeneExpressionMatrix” function. The LSI dimensionality reduction for both scATAC and scRNASeq was computed and combined using the “addCombinedDims” function.

UMAP embedding is calculated using “addUMAP” function and clusters are identified from the reduced dimensions. Marker features were used to annotate clusters. To obtain peaks, “addReproduciblePeakSet” function was used to get insertions from coverage files, call peaks, and merge peaks and this step internally used macs software. Then the “addPeakMatrix” function was called to independently compute counts for each peak per cell. To link peaks to genes, “getPeak2GeneLinks” function was called. The “getMarkerFeatures” function was used to get the differential gene expressions and the differential peaks expressions using the GeneExpressionMatrix and PeakMatrix calculated respectively. Peaks with motifs were annotated using the cisbp motif set using “annotatePeaks.pl” from HOMMER version 4.11 to associate the peaks with nearby genes. Full code used for the analysis is available at https://github.com/SherineAwad/OCT4scRNA-ATAC

eLife Assessment

This manuscript demonstrates that Oct4 overexpression synergizes with Notch inhibition (Rbpj knockout) to promote the conversion of adult murine Müller glia (MG) into bipolar cells. These findings are important as the authors used rigorous genetic lineage tracing (GLAST-CreER; Sun-GFP) to confirm that neurogenesis indeed originates from MGs, addressing a key issue in the field. The single-cell multiomic analyses are convincing, and while functional studies of MG-derived bipolar cells would strengthen the conclusions, they are beyond the scope of this study.

Reviewer #1 (Public review):

Summary:

In this study, Le et al.. aimed to explore whether AAV-mediated overexpression of Oct4 could induce neurogenic competence in adult murine Müller glia, a cell type that, unlike its counterparts in cold-blooded vertebrates, lacks regenerative potential in mammals. The primary goal was to determine whether Oct4 alone, or in combination with Notch signaling inhibition, could drive Müller glia to transdifferentiate into bipolar neurons, offering a potential strategy for retinal regeneration.

The authors demonstrated that Oct4 overexpression alone resulted in the conversion of 5.1% of Müller glia into Otx2+ bipolar-like neurons by five weeks post-injury, compared to 1.1% at two weeks. To further enhance the efficiency of this conversion, they investigated the synergistic effect of Notch signaling inhibition by genetically disrupting Rbpj, a key Notch effector. Under these conditions, the percentage of Müller glia-derived bipolar cells increased significantly to 24.3%, compared to 4.5% in Rbpj-deficient controls without Oct4 overexpression. Similarly, in Notch1/2 double-knockout Müller glia, Oct4 overexpression increased the proportion of GFP+ bipolar cells from 6.6% to 15.8%.

To elucidate the molecular mechanisms driving this reprogramming, the authors performed single-cell RNA sequencing (scRNA-seq) and ATAC-seq, revealing that Oct4 overexpression significantly altered gene regulatory networks. They identified Rfx4, Sox2, and Klf4 as potential mediators of Oct4-induced neurogenic competence, suggesting that Oct4 cooperates with endogenously expressed neurogenic factors to reshape Müller glia identity.

Overall, this study aimed to establish Oct4 overexpression as a novel and efficient strategy to reprogram mammalian Müller glia into retinal neurons, demonstrating both its independent and synergistic effects with Notch pathway inhibition. The findings have important implications for regenerative therapies as they suggest that manipulating pluripotency factors in vivo could unlock the neurogenic potential of Müller glia for treating retinal degenerative diseases.

Strengths:

(1) Novelty: The study provides compelling evidence that Oct4 overexpression alone can induce Müller glia-to-bipolar neuron conversion, challenging the conventional view that mammalian Müller glia lacks neurogenic potential.

(2) Technological Advances: The combination of Muller glia-specific labeling and modifying mouse line, AAV-GFAP promoter-mediated gene expression, single-cell RNA-seq, and ATAC-seq provides a comprehensive mechanistic dissection of glial reprogramming.

(3) Synergistic Effects: The finding that Oct4 overexpression enhances neurogenesis in the absence of Notch signaling introduces a new avenue for retinal repair strategies.

Weaknesses:

(1) In this study, the authors did not perform a comprehensive functional assessment of the bipolar cells derived from Müller glia to confirm their neuronal identity and functionality.

(2) Demonstrating visual recovery in a bipolar cell-deficiency disease model would significantly enhance the translational impact of this work and further validate its therapeutic potential.

Reviewer #2 (Public review):

Summary:

The authors harness single-cell RNAseq data from zebrafish and mice to identify Oct4 as a candidate driver of neurogenesis. They then use adeno-associated virus vectors to show that while Oct4 overexpression alone converts rare adult Müller glia (MG) to bipolar cells, it synergizes with Notch pathway inhibition to cause this neurogenesis (achieved by Cre-mediated knockout of Rbpj floxed allele). Importantly, they genetically lineage-mark adult MG using a GLAST-CreER transgene and a Sun-GFP reporter, so that any non-MG cells that convert can be identified unambiguously. This is crucial because several high-profile papers made erroneous claims using short promoters in the viral delivery vector itself to mark MG, but those promoters are leaky and mark other non-MG cell types, making it impossible to definitively state whether manipulations studied were actually causing neurogenesis, or were merely the result of expression in pre-existing neurons. Once the authors establish Oct4 + RbpjKO synergy they use snRNAseq/ATACseq to identify known and novel transcription factors that could play a role in driving neurogenesis.

Strengths:

The system to mark MG is stringent, so the authors are studying transdifferentiation, not artifactual effects due to leaky viral promoters. The synergy between Oct4 and Notch pathway blockade is notable. The single-cell results add the potential involvement of new players such as Rfx4 in adult-MG-neurogenesis.

Weaknesses:

The existing version is difficult to read due to an unusually high number of text errors (e.g. references to the wrong figure panels etc.). A fuller explanation for the fraction of non-MG cells seen in control scRNAseq assays is required, particularly because the neurogenic trajectory which is enhanced in the Oct4/Rbpj-KO context is also evident in the control retina. Claims regarding the involvement of transcription factors in adult neurogenesis (such as Rfx4) need to be toned down unless they are backed up with functional data. It is possible that such factors are important, but equally, they may have no role or a redundant role, and without functional tests, it's impossible to say one way or the other.

Overall, the authors achieved what they set out to do, and have made new insights into how neurogenesis can be stimulated in MG. Ultimately, a major long-term goal in the field is to replace lost photoreceptors as this is most relevant to many human visual disorders, and while this paper (like all others before it) does not generate rods or cones, it opens new strategies to coax MG to form a related neuronal cell type. Their approach underscores the benefits of using a gold-standard approach for lineage tracing.

Author response:

The following is the authors’ response to the original reviews

Public Reviews:

Reviewer #1 (Public review):

Summary:

In this study, Le et al.. aimed to explore whether AAV-mediated overexpression of Oct4 could induce neurogenic competence in adult murine Müller glia, a cell type that, unlike its counterparts in cold-blooded vertebrates, lacks regenerative potential in mammals. The primary goal was to determine whether Oct4 alone, or in combination with Notch signaling inhibition, could drive Müller glia to transdifferentiate into bipolar neurons, offering a potential strategy for retinal regeneration.

The authors demonstrated that Oct4 overexpression alone resulted in the conversion of 5.1% of Müller glia into Otx2+ bipolar-like neurons by five weeks post-injury, compared to 1.1% at two weeks. To further enhance the efficiency of this conversion, they investigated the synergistic effect of Notch signaling inhibition by genetically disrupting Rbpj, a key Notch effector. Under these conditions, the percentage of Müller gliaderived bipolar cells increased significantly to 24.3%, compared to 4.5% in Rbpjdeficient controls without Oct4 overexpression. Similarly, in Notch1/2 double-knockout Müller glia, Oct4 overexpression increased the proportion of GFP+ bipolar cells from 6.6% to 15.8%.

To elucidate the molecular mechanisms driving this reprogramming, the authors performed single-cell RNA sequencing (scRNA-seq) and ATAC-seq, revealing that Oct4 overexpression significantly altered gene regulatory networks. They identified Rfx4, Sox2, and Klf4 as potential mediators of Oct4-induced neurogenic competence, suggesting that Oct4 cooperates with endogenously expressed neurogenic factors to reshape Müller glia identity.

Overall, this study aimed to establish Oct4 overexpression as a novel and efficient strategy to reprogram mammalian Müller glia into retinal neurons, demonstrating both its independent and synergistic effects with Notch pathway inhibition. The findings have important implications for regenerative therapies as they suggest that manipulating pluripotency factors in vivo could unlock the neurogenic potential of Müller glia for treating retinal degenerative diseases.

Strengths:

(1) Novelty: The study provides compelling evidence that Oct4 overexpression alone can induce Müller glia-to-bipolar neuron conversion, challenging the conventional view that mammalian Müller glia lacks neurogenic potential.

(2) Technological Advances: The combination of Muller glia-specific labeling and modifying mouse line, AAV-GFAP promoter-mediated gene expression, single-cell RNA-seq, and ATAC-seq provides a comprehensive mechanistic dissection of glial reprogramming.

(3) Synergistic Effects: The finding that Oct4 overexpression enhances neurogenesis in the absence of Notch signaling introduces a new avenue for retinal repair strategies.

Weaknesses:

(1) In this study, the authors did not perform a comprehensive functional assessment of the bipolar cells derived from Müller glia to confirm their neuronal identity and functionality.

(2) Demonstrating visual recovery in a bipolar cell-deficiency disease model would significantly enhance the translational impact of this work and further validate its therapeutic potential.

Response: We thank the Reviewer for their evaluation. We agree that functional analysis of Müller glia-derived bipolar cells is indeed important, but is beyond the current scope of the manuscript.

Reviewer #2 (Public review):

Summary:

The authors harness single-cell RNAseq data from zebrafish and mice to identify Oct4 as a candidate driver of neurogenesis. They then use adeno-associated virus vectors to show that while Oct4 overexpression alone converts rare adult Müller glia (MG) to bipolar cells, it synergizes with Notch pathway inhibition to cause this neurogenesis (achieved by Cre-mediated knockout of Rbpj floxed allele). Importantly, they genetically lineage-mark adult MG using a GLAST-CreER transgene and a Sun-GFP reporter, so that any non-MG cells that convert can be identified unambiguously. This is crucial because several high-profile papers made erroneous claims using short promoters in the viral delivery vector itself to mark MG, but those promoters are leaky and mark other non-MG cell types, making it impossible to definitively state whether manipulations studied were actually causing neurogenesis, or were merely the result of expression in pre-existing neurons. Once the authors establish Oct4 + RbpjKO synergy they use snRNAseq/ATACseq to identify known and novel transcription factors that could play a role in driving neurogenesis.

Strengths:

The system to mark MG is stringent, so the authors are studying transdifferentiation, not artifactual effects due to leaky viral promoters. The synergy between Oct4 and Notch pathway blockade is notable. The single-cell results add the potential involvement of new players such as Rfx4 in adult-MG-neurogenesis.

Weaknesses:

The existing version is difficult to read due to an unusually high number of text errors (e.g. references to the wrong figure panels etc.). A fuller explanation for the fraction of non-MG cells seen in control scRNAseq assays is required, particularly because the neurogenic trajectory which is enhanced in the Oct4/Rbpj-KO context is also evident in the control retina. Claims regarding the involvement of transcription factors in adult neurogenesis (such as Rfx4) need to be toned down unless they are backed up with functional data. It is possible that such factors are important, but equally, they may have no role or a redundant role, and without functional tests, it's impossible to say one way or the other.

Overall, the authors achieved what they set out to do, and have made new insights into how neurogenesis can be stimulated in MG. Ultimately, a major long-term goal in the field is to replace lost photoreceptors as this is most relevant to many human visual disorders, and while this paper (like all others before it) does not generate rods or cones, it opens new strategies to coax MG to form a related neuronal cell type. Their approach underscores the benefits of using a gold-standard approach for lineage tracing.

We thank the Reviewer for their evaluation. We have made extensive changes to the manuscript to correct errors and modify discussion as recommended. These are detailed below in our point-by-point responses to specific recommendations to the authors.

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

Minor corrections:

(1) In Figure 1C top GFAP-mCherry panel, two dim GFP + cells have colocalized with Otx2, is it caused by optic imaging thickness or some muller glia cells having the Otx2 expression?

This indeed reflects the effects of optic imaging thickness. Colocalization of Sun1-GFP and Otx2 is not observed when Z-stack images are examined in GlastCreER;Sun1-GFP retinas. This can also be appreciated by the fact that, in cases of apparent overlap of nuclear envelope-targeted Sun1 and Otx2, the sizes of the labeled areas differ. In cases of true expression overlap, such as is seen following Oct4 overexpression, the labeled areas are the same size, or very nearly so.

Whether the Glast-CreERT2 x Rosa26-LSL-Sun1-GFP mouse line has cross-labeling with the Otx2+ bipolar cells, the author should image the mCherry ctrl sample with a thin optical imaging layer with a small pinhole for Z-stack to verify the co-labeling the GFP and Otx2 in mCherry ctrl sample.

Please see above. Since we first described this line (de Melo, et al. 2012), we have examined thousands of sections of GlastCreER;Sun1-GFP retinas, and have yet to see a single GFP-positive neuron. To avoid confusion, however, we have replaced these images with an additional image from a control mCherry-infected GlastCreER;Sun1-GFP retina processed for the same study.

In the middle upper panel, Oct4-mCherry group, the white arrows indicate the GFP colocalized with Otx2 signal, but seems not mCherry positive, by contrast, the neighbor cells have significant mCherry expression but no colocalization with Otx2. The GFAP promoter-Oct4-mCherry may have stopped expression after the Müller Glia cells were converted into Otx2+ bipolar cells, but is there any middle stage in which the Oct4mCherry and Otx2 co-expression? And after Müller glia to Bipolar conversion, why have Glast-CreERT2 driven GFP expressions not suppressed as GFAP promoter driven Oct4-mCherry? Could the author discuss this point?

We observed a significant number of Muller glia-derived cells expressing both Otx2 and weak mCherry signal. GFP expression is driven by the ubiquitous CAG promoter following Cre-dependent excision of a transcriptional stop cassette. We have modified the text to make this point explicit.

(2) In Figure S2b, the mouse is labeled with wild type; I assume it should be the same mouse line as Fig.1. Otherwise, the author should describe the source of the GFP signal.

“Wildtype” in this case refers to GlastCreER;Sun1-GFP controls, which as the Reviewer correctly points out, are not truly wildtype. The genotype of these animals is specified in all figure legends, and is now referred to as “control” rather than “wildtype” in the figures and main text throughout.

In Figure S2k and l, mCherry ctrl panel, the GFP+ cells looked co-labeling with Otx2, so again, is it the thicker optical imaging layer that caused overlapping vertically or the low specific of Müller Glia of the mouse line? Please describe the stars' meaning in Figure S2i,j in the figure legend. There are 2 figures labeled "n" of the quantification data.

This is, again, an example of the thicker optical imaging layer causing apparent overlap. We have previously demonstrated that the Sun1-GFP+ cells do not co-label with Otx2 in GFAP-mCherry AAV-injected control retinas (Le et al., 2022; Fig. 2C). The asterisks (*) indicate mouse-on-mouse vascular staining, which is now clarified in the figure legend. The 2 figures labeled ‘n’ have been relabeled as ‘m’ and ‘n’.

(3) In Figure 2c in the top panel, the Otx2 image was wrong; please replace it with the correct one.

We thank the Reviewer for spotting this error. This is an inadvertent duplication of the single-channel Otx2 staining for mCherry control sample. We have replaced this with the correct image.

(4) In Figure 3a, the Rbpj-cKO mouse line was used, but where was the GFP signal from? Please verify the mouse line you used in your work. The same question is also asked in Figure S3, S4b.

GlastCreER;Rpbjlox/lox;Sun1-GFP were used in Figure 3a. As now specified in the Methods and all figure legends, all mice used in this study carry both the GlastCreER and Sun1-GFP transgenes.

(5) In Figure S4c,d, and 5 wks time point, if the authors quantify the GFP+/Sox2- cells changing, it will be more helpful to understand the percentage of the Müller glia cells conversion to Bipolar cells compared to the Figure 2D, and can be as a supplement to the conclusion Müller to Bipolar conversion rather the Müller proliferation.

Sox2-/GFP+ cells are a measure of Müller glia to bipolar cell conversion that complements that of GFP+/Otx2+ cells. This is now clarified in the text. We also include quantification of Sox2-/GFP+ neurons at 5 weeks post-injury in Fig. S5b.

(6) In Figure S1b,c, there is a large portion of cells that are activated Müller glia after NMDA injury. Did the activated Müller glial cells lose their Müller glial identity? Between the loss of Müller glial identity and neuronal reprogramming, are there any markers that can be used to assess whether Müller glial cells are truly transdifferentiating into neurons rather than remaining in a reactive glial state or an intermediate phase?

Wildtype Müller glia progressively revert to resting state, and by 72 hours post-injury have already lost expression of Klf4 and Myc (Hoang, et al. 2020), a point which is now specifically mentioned in the text. In GlastCreER;Sun1-GFP;Nfia/b/xlox/lox;Rbpjlox/lox Müller glia, reactive MG appear to largely convert to bipolar and amacrine-like cells, and it remains unclear if they eventually revert to a resting state (Le, et al. 2024).

Reviewer #2 (Recommendations for the authors):

This work demonstrates that Oct4 (Pou5f3) can induce neurogenesis in murine Müller glia (MG). Le et al start by showing that murine and zebrafish MG lack expression of Oct4 (Pou5f3) and its target Nanog. To assess the effect of Oct4 they first label adult MG with Sun1-GFP using tamoxifen-treated GlastCreER;Sun1-GFP mice, then later transduce in vivo with AAV vectors expressing mCherry alone or Oct4 + mCherry. Subsequently, they damage the retina with NMDA and assess the effects several weeks later. In Oct4+ cells at 2 weeks there is rare induction of the neural determinant Ascl1, down-regulation of the MG marker Sox2, induction of bipolar markers (Otx2, Scgn,Cabp5) but not amacrine (HuC/D) or rod (Nrl) markers. Combining Oct4 with

Notch inhibition (deleting floxed Rbpj) synergistically increases bipolar cell induction, with Otx2 staining rising to >20% of GFP-marked cells, and cells losing MG identify (loss of Sox2/9). EdU labeling was negligible suggesting direct trans-differentiation. Similar synergy was seen upon combining Oct4 expression with Notch1/2 double gene knockout. Attempts to combine Oct4 with Nfia, Nfib, and Nfix loss were unsuccessful as the GFAP promoter driving Oct4 in MG seems to require these three related transcription factors. scRNAseq confirmed the Oct4-overexpression/Rbpj-KO-driven increase in bipolar cells and decrease in MG cells and revealed that these manipulations may enhance bipolar cell genesis by repressing genes that define quiescent MG and enhancing expression of genes that define reactive MG and neurogenic cells. Finally, multiomic snRNA/scATAC-seq data was performed to assess the effect of Oct2 in wt or Rbpj null MG. This approach revealed that, as anticipated, more genes were up and down-regulated in the context of both manipulations vs Oct4 OE alone. Moreover, Oct4 and Rbpj KO reduced chromatin accessibility at target motifs for transcription factors involved in MG identify/quiescence, while MGPCs showed elevated accessibility for neurogenic factors. The combination of Oct4 OE and Rbpj KO induces accessibility at various interesting TF sites that may contribute to the synergistic neurogenesis, including Rfx4, Klf4, Insm1, and others.

This is an interesting paper that adds to the growing literature on how neurogenesis can be induced in mammalian MG. The focus on Oct4 is interesting and the synergistic effects are striking and analyzed in some detail with scRNAseq and multiomic snRNA/scATACseq. The latter results provide useful new insight into transcriptional programs that may be critical in driving neurogenesis. Functional insight into these new candidates is not explored in this manuscript, but that's beyond the scope of the current work and forms the basis for new studies. There are some overreaching statements in the Discussion that need to be toned down, but apart from that and a long list of textual errors that need to be fixed, this paper is a valuable contribution to the field.

Major comments

There are numerous textual errors (some, but not all, examples are detailed in minor comments). It was difficult to follow this paper given the unusually high number of textual errors and the abbreviated legends. Greater attention should be paid to harmonizing the text with the figures and ensuring that the legends are correct and complete.

The manuscript has been proofread carefully and errors corrected.

The opening section of the scRNAseq data should outline briefly why sorting for GFP labeled cells purifies a significant fraction of non-MG cell types, despite the earlier claim, (which agrees with other publications), that GLAST-CreER transgene expression is highly specific to MG. Presumably, it mainly/totally reflects the co-purification of cells, cell fragments, and/or cell-free mRNA from other lineages. Is it also possible that a fraction (however small) of these cells reflect low-level spurious/temporary activation of GLAST-CreER expression in non-MG? The "contamination" is present despite the addition of the GFP sequence to the reference genome (as explained in Methods). They mention: "a clear differentiation trajectory connecting Muller glia, neurogenic Muller gliaderived progenitor cells (MGPCs), and differentiating amacrine and bipolar cells (Fig. 3b)". However, the same trajectory is evident in control mCherry samples, so one could argue that this trajectory is active in normal retina at some low rate, but that would/should equate to rare sun-GFP+ non-MG in controls. Are there any such cells, even extremely rarely, or is it truly 0%? At any rate, the authors need to raise these concerns and offer some explanation(s) at the start of their scRNAseq Results section. If there are really no such sun-GFP+ cells, the authors should comment on the presence of the apparent inactive trajectory in the Discussion.

Since we first described this line (de Melo, et al. 2012), we have examined thousands of sections of GlastCreER;Sun1-GFP retinas, and have yet to see a single GFP-positive neuron. We have also previously shown (Hoang, et al. 2020) that FACSbased isolation of GFP-positive cells from GlastCreER;Sun1-GFP yields a roughly thirty-fold enrichment of Muller glia, implying the presence of small numbers of contaminating neurons. We thereby conclude that the presence of small numbers of neurons (rods, cones, bipolar, and amacrine cells) in the control GlastCreER;Sun1-GFP represents contamination rather than low levels of glia-to-neuron conversion, particularly since we are unable to detect the expression of genes such as neurogenic bHLH factors or immature photoreceptor precursor-specific factors such as Prdm1 that indicate the presence of intermediate cell states. This is now addressed in the Results section related to both Figures 3 and 4.

Discussion:

In reference to other strategies to induce neurogenesis the authors make the claim that Oct4 is fundamentally different: "In these cases, Müller glia broadly upregulate proneural genes and/or downregulate Notch signaling. Oct4 instead induces expression of the neurogenic transcription factor Rfx4, which is not expressed in developing retina. It is likely that activation of this parallel pathway to neurogenic competence in part accounts for synergistic induction of neurogenesis seen in Rbpj-deficient Müller glia". First, all these strategies, including Oct4, seem to activate bHLH factors, so they have that in common and the authors should note that overlap. More seriously, without functional tests (e.g. KO Rfx4) the authors need to dial back the over-reaching statement that Rfx4 is the fundamental mechanism driving the Oct4 effect. They can certainly suggest that this is one possibility, but equally, Rfx4 may have very little or no effect on neurogenesis, or it could act redundantly with some of the other factors the authors uncovered. It's impossible to know without functional data, so they either need to add the functional data, or hold back on the strong one-sided and overreaching claim.

Since both Rfx4 expression and motif accessibility are selectively observed following Oct4 overexpression, and Rfx4 also has known neurogenic activity, we stand by our conclusion that it is a particularly strong candidate for mediating the neurogenic effects of Oct4 overexpression. However, the Reviewer is correct that in the absence of functional data, speculation about its function should be qualified. We have done this in the revised manuscript.

Minor comments

This sentence in the Results is confusing: "While expression of neurogenic bHLH factors driven by the Gfap promoter was rapidly silenced in Muller glia and activated in amacrine and retinal ganglion cells, Gfap-Oct4-mCherry remained selectively expressed in Muller glia but did not induce detectable levels of Muller glia-derived neurogenesis in the uninjured retina (Le et al., 2022)". The cited reference is at the end so it sounds like the Oct4 assay was performed in Le et al 2022, and there is no reference to a Figure for the Oct4 data in the current paper.

As stated here, in Le, et al. 2022, we did not observe any conversion of Sun1-GFP-positive Muller glia to neurons in the absence of injury. In the current study, we instead test whether NMDA-induced excitotoxicity induced glia to neuron conversion in Muller glia overexpressing Oct4. This is now made clear in the revised text.

There are many errors and omissions regarding Figure S2:

Figure S2a, b legend, and panels do not match. 2a should be a schematic of the strategy to label MG with Sun1-GFP using GLAST-Cre and a floxed Sun1-GFP allele, but that's missing and instead, the current 2a is a schematic of AAV vectors. It seems that the current 2b legend may describe the combination of the current 2a and 2b panels.

This has been corrected.

Figure S2: Asterisks label certain stained elements in the Oct4 labeled panels, but there is no explanation in the legend. Are these meant to indicate non-specific staining? If so, what is the evidence that the signal is non-specific?

These asterisks represent non-specific mouse-on-mouse vascular staining observed with the mouse monoclonal anti-Oct4 used in this study. This is now indicated in the figure legend.

The text refers to Ascl1 staining in Figure S2e,f, but it's S2g,h.

This has been corrected.

Re this: "While Sun1-GFP-positive cells infected with Oct4-mCherry mostly express the Muller glial marker Sox2 (Fig. S2a,b), from 2 weeks post-injury onwards a subset of GFP positive cells did not show detectable Sox2 expression (Fig. S2b, yellow arrows)". Figure S2a, b are schematic diagrams, not immunofluorescence. They probably mean Figure S2c, d.

This has been corrected.

Fig S2m is mislabeled "n".

This has been corrected.

There are probably other errors with this figure, but I mostly gave up at this point. The authors should go through the paper to find and correct any additional mistakes/omissions in the text and legends.

The manuscript has been carefully proofread and errors corrected.

The figure panels are not always mentioned in the order that they appear. There are many examples.

Figure panels are now mentioned in the order that they appear.

Several schematics use "d-18-14" to indicate "day -18 to -14". The former is at first uninterpretable or at best unclear (could mean day -18 to day 14), perhaps d -18 to -14, or d -18:-14 would be clearer.

This has been corrected.

Re: "AAV-infected wildtype Muller glia could be readily identified by selective expression of Oct4 (Fig. 4e). Wildtype Oct4-expressing Muller glia give rise to both small numbers of neurogenic MGPCs (Fig. 4b),". Figure 4E is labeled Pou5f1, but it would be helpful to avoid confusion by also indicating on the figure that Pou5f1 = Oct4; and Fig 4b does not indicate neurogenic MGPCs (perhaps they mean 4c).

This has been corrected.

Some parts of the Results are written in the present tense and should be in the past tense (for guidance: https://www.nature.com/scitable/topicpage/effective-writing13815989/).

Past tense is now used throughout.

Pit1 (Pou1f1) is referred to as a "close variant" of Oct4/Pou4f5, but this is unclear (e.g. variant could mean a splice variant from the same locus) and the term "paralogue" should be used.

“Paralogue” is now used in this context.

Re: "Infection with Oct4-mCherry vector induced both Oct4 (Fig. S5e) and Ascl1 (Fig. S5d) expression in Notch1/2-deficient Müller glia." Supplementary image 5d is the one depicting Oct4 and 5e is the one showing Ascl1. However, the reference is reversed.

This has been corrected.

Version 2

Abstract

Abstract

Retinal Müller glia in cold-blooded vertebrates can reprogram into neurogenic progenitors to replace neurons lost to injury, but mammals lack this ability. While recent studies have shown that transgenic overexpression of neurogenic bHLH factors and glial-specific disruption of NFI family transcription factors and Notch signaling induce neurogenic competence in mammalian Müller glia, induction of neurogenesis in wild-type glia has thus far proven elusive. Here, we report that viral-mediated overexpression of the pluripotency factor Oct4 (Pou5f1) induces transdifferentiation of mouse Müller glia into bipolar neurons, and synergistically stimulates glial-derived neurogenesis in parallel with Notch loss of function. Single cell multiomic analysis shows that Oct4 overexpression leads to widespread changes in gene expression and chromatin accessibility, inducing activity of both the neurogenic transcription factor Rfx4 and the Yamanaka factors Sox2 and Klf4. This study demonstrates that viral-mediated overexpression of Oct4 induces neurogenic competence in retinal Müller glia, identifying mechanisms that could be used in cell-based therapies for treating retinal dystrophies.

Introduction

The homeodomain factor Oct4 (Pou5f1) is a central component of transcriptional regulatory networks that maintain pluripotency in embryonic stem (ES) cells. Along with Sox2, Klf4, and Myc, Oct4 is one of the Yamanaka factors sufficient to induce the formation of induced pluripotent stem cells (iPSCs) from virtually all somatic cell types (Takahashi and Yamanaka, 2006). In contrast to these other factors, Oct4 is selectively expressed in ES cells and essentially absent from somatic cells, including adult stem cells from every tissue examined. The two clear exceptions are found in gonadal stem cells (Takashima et al., 2013; Uhlen et al., 2010) and during neural crest formation, where Oct4 expression is essential for the ability of these cells to generate diverse cell types (Morrison and Brickman, 2006; Zalc et al., 2021). This has raised the question of whether Oct4 overexpression in vivo, alone or in combination with other pluripotency factors, may confer some level of multipotency on somatic cells in the central nervous system.

Sustained overexpression of the Yamanaka factors Oct4, Sox2, Klf4, and Myc induces teratoma formation in many tissues (Abad et al., 2013; Ohnishi et al., 2014). Transient or low-level Yamanaka factor overexpression, in contrast, has been reported to induce cellular rejuvenation in both neurons and other cell types, although this occurs without clear induction of proliferation or conferring multipotency (Ocampo et al., 2016; C. Wang et al., 2021). However, several studies have directly investigated whether viral-mediated overexpression in CNS neural stem or glial cells of Oct4 alone, or in combination with subsets of other pluripotency factors, can induce multipotency and neurogenesis. Results from these studies are mixed, with Oct4 overexpression in neural stem cells reported to either not induce neurogenesis (Asadi et al., 2015; Dehghan et al., 2015; Sim et al., 2011), to promote formation of oligodendrocyte progenitors (OPCs) (Yu et al., 2021), or to promote remyelination in committed OPCs (Dehghan et al., 2016). Two studies reported that viral-mediated Oct4 overexpression, either alone (Niu et al., 2013) or in combination with other Yamanaka factors (Gao et al., 2016), could directly convert reactive astrocytes into neurons. However, these studies used the GFAP minipromoter to drive gene expression, which has been reported to show insert-dependent ectopic insert-dependent silencing in glia and activation in neurons (L.-L. Wang et al., 2021). This limitation makes these findings difficult to interpret, as additional methods of validating cell lineage were not used.

Two previous studies have reported injury-dependent induction of Oct4 in retinal Müller glial cells in zebrafish and mice. In zebrafish, where injury induces Müller glia to undergo conversion to neurogenic progenitors (Wan and Goldman, 2016), it was reported that Oct4 is essential for glial reprogramming to occur (Sharma et al., 2019). In mice, where neurogenic competence is rapidly and actively suppressed following injury, it was reported that Oct4 expression is transiently induced, but fully repressed within 24 hours following injury (Reyes-Aguirre and Lamas, 2016). This raises the possibility that Oct4 expression may be sufficient to confer neurogenic competence on retinal Müller glial cells. In this study, we assessed whether AAV-mediated overexpression of Oct4 could induce reprogramming of adult mouse Müller glia. While we did not detect any expression of Oct4 in either zebrafish or mouse Müller glia following injury, we used both genetic cell lineage and single-cell omics analyses to unambiguously demonstrate that viral-mediated overexpression of Oct4 selectively induces the formation of bipolar neurons from control GlastCreER;Sun1-GFP mouse Müller glia without inducing substantial levels of proliferation. We further showed that Oct4 overexpression synergistically enhances existing levels of Müller glia-derived neurogenesis in the absence of Notch signaling. Single-cell multiomic analysis demonstrated that Oct4 cooperates with Sox2 and Klf4, which are constitutively expressed in Müller glia, to broadly alter the gene regulatory landscape and induce expression of genes such as the neurogenic transcription factor Rfx4. This study demonstrates that substantial levels of in situ glia-to-neuron reprogramming in mammalian retina can be induced by viral overexpression of a Yamanaka factor.

Results

Oct4 and Nanog are not detectably expressed in Müller glia or Müller glia-derived progenitor cells in either zebrafish or mice

To examine the expression of Oct4 and other pluripotency-associated genes in Müller glia, we first analyzed previously published gene expression data from zebrafish and mouse retina (Hoang et al., 2020; Le et al., 2024). We first analyzed both single-cell RNA-Seq data obtained from wildtype retinas and bulk RNA-Seq data obtained from FACS-isolated Müller glia carrying glial-specific GFP reporter lines. Consistent with previous findings (Nelson et al., 2011; Surzenko et al., 2013), in mice we detected expression of Sox2 and Klf4 in resting and both NMDA-treated and light-damaged Müller glia, along with injury-induced expression of Myc, which is lost as activated glia progressively return to a resting state by 72 hours post-injury.

Likewise, in zebrafish we observed expression of sox2 and myca/b (but not klf4) in both resting and injured Müller glia (Fig. S1a,b). However, we did not detect expression of either Oct4 or its direct target Nanog in either species in any treatment condition (Loh et al., 2006). Bulk RNA-Seq data obtained from purified Müller glia detected very low levels of Oct4 (pou5f3), but not Nanog, expression in some injured zebrafish Muller glia, but neither Oct4 nor Nanog was detected in either resting or injured mouse Müller glia (Fig. S1a,b). In addition, we did not observe any induction of Oct4 or Nanog in snRNA-Seq data obtained from mouse Müller glia following selective loss of function of Rbpj alone or in combination with Nfia/b/x (Le et al., 2024) (Fig. S1c). These results indicate that neither NMDA excitotoxicity, light damage, or induction of neurogenic competence by loss of function of Nfia/b/x and/or Rbpj induces detectable levels of Oct4 or Nanog in neurogenic mammalian Müller glia.

AAV-mediated overexpression of Oct4 induces conversion of control GlastCreER;Sun1-GFP mouse Müller glia to bipolar-like cells

To determine whether Oct4 overexpression could reprogram mammalian Müller glia into a neurogenic state, we administered intraperitoneal tamoxifen injections into P21 GlastCreER;Sun1-GFP mice, thereby irreversibly labeling Müller glia and their progeny with nuclear envelope-targeted Sun1-GFP under control of the ubiquitous CAG promoter (de Melo et al., 2012; Hoang et al., 2020) (Fig. 1a, Fig. S2a). One week following the final tamoxifen dose, we performed intravitreal injections of either Gfap-Oct4-mCherry or Gfap-mCherry control 7m8 AAV (Fig. 1b, Fig. S2b). While expression of other neurogenic bHLH factors driven by the GFAP promoter were rapidly silenced in Müller glia and activated in amacrine and retinal ganglion cells, Gfap-Oct4-mCherry remained selectively expressed in Müller glia. However, Oct4 expression alone did not induce detectable levels of Müller glia-derived neurogenesis in the uninjured retina (Le et al., 2022).

AAV-mediated overexpression of Oct4 induces neurogenesis in control GlastCreER;Sun1-GFP Müller glia following NMDA-induced excitotoxicity.

(A) Schematic of the transgenic construct used to specifically label Müller glia with Sun1-GFP expression. (B) Schematic of the GFAP AAV constructs and experimental workflow. (C) Representative images of retinas immunolabeled for GFP, mCherry, Otx2, and HuC/D. White arrowheads indicate GFP-positive Müller glia-derived neurons expressing neuronal markers Otx2 or HuC/D. (D) Quantification of mean percentage ± SD of GFP-positive Müller glia-derived neurons expressing either Otx2 or HuC/D. Significance was determined via two-way ANOVA with Tukey’s multiple comparison test: ****p < 0.0001. Each data point was calculated from an individual retina. TAM, tamoxifen; ONL, outer nuclear layer; INL, inner nuclear layer; GCL, ganglion cell layer. Scale bar = 50μm.

Since injury is often required to promote Müller glia-derived neurogenesis following Ascl1 overexpression or loss of function of Nfia/b/x (Hoang et al., 2020; Jorstad et al., 2017), we hypothesized that this might also be so for Müller glia overexpressing Oct4. To test this, we injected N-methyl-d-aspartate (NMDA) one week post-viral injection. We then administered 5-ethynyl-2’-deoxyuridine (EdU) to label proliferating cells. Retinas were harvested and analyzed at 1, 2, and 5 weeks post-injury (Fig. S2b, Fig. 1b). Immunostaining analysis revealed enriched mCherry expression in GlastCreER;Sun1-GFP Müller glia in both control mCherry and Oct4-mCherry infected retinas at all timepoints (Fig. 1c, Fig. S2c-l). In addition, Oct4 expression was detected in GFP-positive Müller glia in Oct4-mCherry infected retinas but not in mCherry infected GlastCreER;Sun1-GFP retina (Fig. S2c,d, Fig. S3a). While most Sun1-GFP-positive cells infected with Oct4-mCherry expressed the Müller glial marker Sox2 (Fig. S2e,f, S3b), a subset of GFP-positive cells lost Sox2 expression from 2 weeks post-injury onward (Fig. S2f, Fig. S3b, yellow arrows), which is consistent with glia-to-neuron conversion.

We observed selective expression of the neurogenic bHLH factor Ascl1 in Oct4-mCherry infected GFP-positive Müller glia at 1 week post-injury (Fig. S2g), with the number of Ascl1-positive cells declining somewhat by 2 and 5 weeks post-injury (Fig. S2h, Fig. S3c). While at 1 week post-injury, we did not observe any significant Müller glia-derived neurogenesis (Fig. S2i,m), by 2 weeks post-injury we observed 1.1% of GFP-positive cells co-localized with the bipolar cell marker Otx2 in Oct4-mCherry infected retinas (Fig. S2j,m), and this increased to 5.1% of GFP-positive cells at 5 weeks post-injury (Fig. 1c,d). We found that Müller glia-derived Otx2+ bipolar-like cells lost mCherry reporter expression, likely due to reduced activity of the Gfap-Oct4-mCherry construct in neurons (Le et al., 2022). Immunostaining with additional cell-specific markers revealed that a subset of these GFP-positive cells expressed the cone bipolar markers Scgn and Cabp5 (Fig. S3d,e), but not the rod and cone ON bipolar cell marker Isl1 or the rod marker Nrl (Fig. S3f,g). Furthermore, we did not observe expression of the amacrine cell marker HuC/D in GFP-positive cells (Fig. 1c,d), which was observed in neurons generated from neurogenic mouse Müller glia deficient in either Rbpj or Nfia/b/x (Hoang et al., 2020; Le et al., 2024). No EdU/GFP-positive cells were detected at either 1 week (Fig. S2k) or 5 weeks post-injury (Fig. S3h). However, at 2 weeks post-injury, we observed a very limited number of GFP-positive cells incorporating EdU in retinas infected with both mCherry control and Oct4-mCherry, with modest EdU incorporation but significantly increased by Oct4 overexpression (Fig. S2l,n).

These findings indicate that Oct4 overexpression induces a very low level of proliferation in Müller glia in the second week following NMDA injury. However, the absence of EdU/Otx2/GFP-positive cells at both 2 and 5 weeks post-injury suggests that bipolar cell generation at this stage was, at least in part, driven by direct transdifferentiation.

Oct4 overexpression synergistically enhances glial-derived neurogenesis induced by genetic disruption of Notch signaling

In our previous study, we demonstrated that disrupting Notch signaling in adult mouse Müller glia – either by selective deletion of the transcriptional mediator Rbpj or by combined disruption of Notch1 and Notch2 receptors – induces transdifferentiation into bipolar and amacrine-like cells (Le et al. 2024). To further determine the effects of Oct4 overexpression in the absence of Notch signaling, we used GlastCreER;Rbpjlox/lox;Sun1-GFP mice to genetically disrupt Rbpj function (Fig. 2a, Fig. S4a). We then performed intravitreal injection of either control or Oct4 overexpression constructs followed by NMDA and EdU injection as performed for GlastCreER;Sun1-GFP mice. Retinas were collected at 1, 2, and 5 weeks following NMDA treatment (Fig. 2b, Fig. S4b). Immunostaining confirmed expression of Oct4 protein in GFP-positive Müller glia of Oct4-mCherry infected retinas, but not in age-matched mCherry-infected controls (Fig. S4c,d, Fig. S5a). We observed that many GFP-positive cells in Oct4-mCherry-infected retinas did not express the Müller glial markers Sox2 (Fig.S4e,f) or Sox9 (Fig. S5b, yellow arrows), suggesting that they have lost their glial identity.

Oct4 overexpression enhances neurogenesis in Rbpj-deficient GlastCreER;Rbpjlox/lox;Sun1-GFP Müller glia.

(A) Schematic of the transgenic constructs used to induce loss of function of Rbpj specifically in Müller glia. (B) Schematic of the GFAP AAV constructs and experimental workflow. (C) Representative images of retinas immunolabeled for GFP, mCherry, Otx2, and HuC/D. White arrowheads indicate GFP-positive Müller glia-derived neurons expressing neuronal markers Otx2 or HuC/D. (D) Quantification of mean percentage ± SD of GFP-positive Müller glia-derived neurons expressing either Otx2 or HuC/D. Significance was determined via two-way ANOVA with Tukey’s multiple comparison test: ****p < 0.0001. TAM, tamoxifen; ONL, outer nuclear layer; INL, inner nuclear layer; GCL, ganglion cell layer. Scale bar = 50μm.

At 1 and 2 weeks post retinal injury, we observed selective expression of the neurogenic bHLH factor Ascl1 in both control mCherry and Oct4-mCherry infected GFP-positive Rbpj-deficient Müller glia (Fig. S4g,h). Time course analysis showed that Müller glia-derived cells expressing the bipolar cell marker Otx2 appeared as early as 1 week following NMDA injury (Fig. S4i-k). However, we did not observe a significant change in the proportion of GFP-positive Rbpj-deficient Müller glia expressing the amacrine cell marker HuC/D. By five weeks after NMDA injury, we observed 4.5% of GFP-positive cells in retinas of mCherry-infected controls expressed Otx2, while 0.7% expressed HuC/D in Oct4-mCherry infected retinas (Fig. 2c,d).

These levels of neurogenesis are consistent with previous findings in uninfected Rbpj-deficient Müller glia (Le et al., 2024). In contrast, we observed 24.3% of GFP-positive cells expressing Otx2 in Oct4-mCherry-infected retinas, representing a significant (p < 0.0001) increase relative to mCherry control. Unlike in control (GlastCreER;Sun1-GFP) retinas, Oct4-expressing, Rbpj-deficient Müller glia showed no EdU incorporation at either 1, 2 or 5 weeks following infection (Fig. S4l,m, Fig. S5c). EdU labeling instead colocalized with the microglia marker Iba1. The absence of Müller glia proliferation indicates that the observed neurogenesis is due largely to direct transdifferentiation of Müller glia into retinal neuron-like cells, rather than through dedifferentiation to a proliferative neuronal progenitor-like state.

We have previously shown that Müller glia-specific Notch1/2 double loss of function mutants phenocopy neurogenesis seen in Rbpj mutants (Le et al., 2024). We then examined the effect of Oct4 overexpression on GlastCreER;Notch1lox/lox;Notch2lox/lox;Sun1-GFP mice using the same protocol previously described for control and Rbpj-deficient Müller glia (Fig. S6a,b). We observed 15.8% of GFP+ Müller glia colocalized with Otx2 in Oct4-mCherry-infected retinas compared to 6.6% GFP/Otx2+ cells in samples infected with control mCherry vector (Fig. S6c,d). As expected, infection with Oct4-mCherry vector induced both Oct4 (Fig. S6e) and Ascl1 (Fig. S6f) expression in Notch1/2-deficient Müller glia.

Finally, we tested whether Oct4 overexpression could enhance Müller glia-derived neurogenesis induced by loss of function of the transcription factors Nfia, Nfib, and Nfix (Hoang et al., 2020). To test this, we injected GlastCreER;Nfialox/lox;Nfiblox/lox;Nfixlox/lox;Sun1-GFP mice with both mCherry and Oct4-mCherry vectors as previously described, then analyzed the retinas 5 weeks after NMDA injury (Fig. S7a,b). We observed limited coexpression of mCherry and GFP following infection with either construct (Fig. S7c), indicating that NFI factors may be required for appropriate glial-specific expression of the GFAP minipromoter construct, as suggested by previous findings in astrocytes (Cebolla and Vallejo, 2006).

Oct4 overexpression promotes expression of reactive glial markers and neurogenic bHLH factors in Rbpj-deficient Müller glia

To gain insight into the mechanism by which Oct4 stimulates neurogenesis in Rbpj-deficient Müller glia, we conducted single-cell (sc)RNA-Seq analysis of GFP-positive cells FACS-isolated from retinas of GlastCreER;Rbpjlox/lox;Sun1-GFP mice infected with either GFAP-mCherry control or GFAP-Oct4-mCherry virus. Mice were tamoxifen- and NMDA-treated as described in Figure 2, and GFP-positive cells were isolated at 8 weeks following NMDA treatment (Fig. 3a). We profiled 12,143 GFP+ cells from control-infected and 14,612 GFP+ cells from Oct4-mCherry-infected retinas (Fig. 3b).

ScRNA-Seq analysis of Müller glia and Müller glia-derived neurons from Rbpj-deficient GlastCreER;Sun1-GFP;Rbpjlox/loxMüller glia following Oct4 overexpression.

(A) Schematic of the scRNA-seq experimental pipeline. (B) UMAP plot showing the clustering of GFP-positive cells from Rbpj-deficient GlastCreER;Rbpjlox/lox;Sun1-GFP Müller glia retinas infected with GFAP-mCherry and GFAP-Oct4-mCherry AAV constructs. (C) Stacked bar plots showing the proportion of cells in each cluster across two sample groups. (D) Feature plots highlighting the cluster of Müller glia (Hes5), neurogenic MGPCs (Neurog2, Insm1, Ascl1), bipolar cells (Otx2), and amacrine cells (Elavl3). (E) Heatmap showing the expression of top differentially expressed genes (DEGs) from the Rbpj-deficient GlastCreER;Rbpjlox/lox;Sun1-GFP Müller glia cell cluster from retinas infected with GFAP-mCherry and GFAP-Oct4-mCherry AAV constructs. (F) Heatmap showing the expression of top differentially expressed genes (DEGs) for MGPC cell cluster from retinas infected with GFAP-mCherry and GFAP-Oct4-mCherry AAV constructs.

In both samples, we observed a clear differentiation trajectory connecting Müller glia, neurogenic Müller glia-derived progenitor cells (MGPCs), and differentiating amacrine and bipolar cells (Fig. 3b). Oct4-overexpressing samples consistently show substantially higher fractions of bipolar cells (20.5 % vs. 2.7 %), and a smaller fraction of Müller glia (71.4 % vs. 88.2 %), consistent with immunohistochemical data (Fig. 3c). Small numbers of contaminating rod and cone photoreceptors were also observed, as previously reported (Hoang et al., 2020), but no immature photoreceptor-like cells were observed, indicating that these represent contaminating mature cells that were not excluded by FACS analysis.

Selective molecular markers could readily distinguish Müller glia (Glul, Apoe, Vim, Clu, Rlbp1) from neurogenic MGPCs (Neurog2, Ascl1, Sox4, Hes6, Dll1), and clearly identify differentiating bipolar (Otx2, Pcp2, Pcp4, Car10, Gabrb3) and amacrine-like cells (Nrxn1, Tfap2b, Nrg1, Snap25) (Fig. 3d-f, Fig. S8a-c, Table ST1). In both MGs and MGPCs, we observed substantial differences in gene expression between samples infected with control and Oct4-mCherry samples. Oct4-infected Müller glia selectively downregulated genes specific to resting Müller glia (Car2, Dkk3, Glul, Col23a1, Abca8a), and upregulated genes enriched in reactive glia (Gfap, Socs3, Nupr1, Mt1/2/3, Fos, Jun, Klf6) (Fig. 3e-f, Fig. 8b-c, Table ST1). In MGPCs, Oct4 overexpression also downregulated both transcription factors (Lhx2, Hopx) and other genes (Apoe, Dio1, Il33, Espn) that are enriched in resting Müller glia, and upregulated many other genes that are selectively expressed in activated glia (Mt1/2/3, Fos, Jun, Nupr1, Socs3). In addition, Oct4-infected MGPCs expressed consistently higher levels of neurogenic bHLH factors (Ascl1, Neurog2, Hes6), bipolar cell specification (Otx2), or genes broadly associated with neurogenesis (Insm1, Dll3, Mybl1). Oct4-infected MGPCs also expressed genes (Gadd45a, Btg2) enriched in late-stage neurogenic progenitors (Clark et al., 2019) (Fig. 3e-f, Table ST1). Together, these findings imply that Oct4 enhances bipolar cell formation in Rbpj-deficient Müller glia potentially by repressing expression of genes such as Lhx2 that promote quiescence (de Melo et al., 2018, 2016), increasing expression of genes that broadly promote neurogenesis, and selectively increasing expression of the bipolar cell-promoting factor Otx2 (Chan et al., 2020; Wang et al., 2014).

Single-cell multiomic analysis identifies targets of Oct4 in control and Rbpj-deficient Müller glia

To further investigate the mechanism by which Oct4 promotes neurogenesis, we conducted simultaneous single-nucleus (sn)RNA-Seq and ATAC-Seq on GFP-positive cells from both control and Rbpj-deficient Müller glia infected with either control or Oct4-mCherry vectors at 8 weeks following NMDA injection (Fig. 4a). UMAP analysis of integrated snRNA/ATAC-Seq profiles was then used to identify cell clusters. Müller glia from control and Rbpj-deficient animals form two separate Müller glia clusters (Fig. 4b,c). Control Oct4-expressing Müller glia gave rise to both small numbers of neurogenic MGPCs (Fig. 4d), in line with the small numbers of Ascl1-positive cells observed (Fig. S2g,h, Fig. S3b), as well as mature bipolar cells, consistent with histological findings (Fig. 1c,d). Oct4 overexpression substantially enhances bipolar cell generation from Rbpj-deficient MGPC (Fig. 4d, Fig. S9a-c). Small numbers of rod and cone photoreceptors were also detected in this analysis, but their proportions did not substantially change between treatment conditions and clearly defined immature transitional states were identified, and therefore likely represent contaminating native cells.

Integrated snRNA/scATAC-Seq analysis of control GlastCreER;Sun1-GFP and Rbpj-deficient GlastCreER;Rbpjlox/lox;Sun1-GFP Müller glia following Oct4 overexpression.

(A) Schematic of the multiomic scRNA/ATAC-seq experimental pipeline. (B) UMAP plot of multiomic datasets showing the clustering of GFP+ cells from control GlastCreER;Sun1-GFP and Rbpj-deficient GlastCreER;Rbpjlox/lox;Sun1-GFP Müller glia from GFAP-mCherry and GFAP-Oct4-mCherry AAV infected retinas. (C) UMAP plot showing the identity of cell clusters determined by marker gene expression. (D) Stacked bar plots represent the proportion of cells in each cluster across different sample groups. (E) Feature plots highlighting the cluster of Müller glia (Hes1), neurogenic MGPCs (Neurog2, Sox4), bipolar cells (Otx2), Pou5f1- and Rfx4-expressing cells. (F) Heatmap showing expression of DEGs along the neurogenesis trajectory. (G) Heatmap showing differential motif activity along the neurogenesis trajectory. (H) Heatmaps of DEGs differentially expressed between Oct4 and mCherry control samples in the control GlastCreER;Sun1-GFP and Rbpj-deficient GlastCreER;Rbpjlox/lox;Sun1-GFP Müller glia clusters. Venn diagram showing unique and common DEGs between Oct4 and mCherry control in the control GlastCreER;Sun1-GFP and Rbpj-deficient GlastCreER;Rbpjlox/lox;Sun1-GFP Müller glia. (I) Increased chromatin accessibility regions associated with the Oct4 and Rfx4 loci observed following Oct4 overexpression in both control GlastCreER;Sun1-GFP and Rbpj-deficient GlastCreER;Rbpjlox/lox;Sun1-GFP Müller glia. (J) Top-ranked enriched motifs in chromatin regions showing increased accessibility following Oct4 overexpression in both control GlastCreER;Sun1-GFP and Rbpj-deficient GlastCreER;Sun1-GFP;Rbpjlox/lox Müller glia cell clusters.

Oct4 overexpression leads to significant changes in gene expression in both control and Rbpj-deficient Müller glia. The transition from resting to neurogenic MGPCs to bipolar cells is associated with downregulation of genes that are specifically expressed in resting Müller glia, including Aqp4, Hes1, Tcf7l2, and Lhx2 (Fig. 4e,f, Fig. S9a-c, Table ST2). Similarly, motif analysis showed that chromatin accessibility is reduced for target motifs for transcription factors that maintain glial quiescence such as Nfia/b/x and Lhx2 (Fig. 4g, Table ST2). MGPCs likewise show increased accessibility at Stat1/3 motifs, as well as target motifs for neurogenic transcription factors such as Ascl1, Neurog2, and Neurod1 (Fig. 4g), consistent with previous studies of MGPCs generated from Nfia/b/x and Rbpj-deficient Müller glia (Hoang et al., 2020; Le et al., 2024).

Oct4 overexpression in both control and Rbpj-deficient Müller glia clusters respectively induces upregulation of 534 and 1657 genes, while respectively downregulating 439 and 824 genes (Fig. 4h). A total of 156 genes are upregulated in both control and Rbpj-deficient Müller glia, while 63 genes are downregulated. In both control and Rbpj-deficient Müller glia, Oct4 upregulates multiple classes of ion channels, including mechanosensitive (Piezo2), TRP (Trpm1, Trpm3), and chloride channels (Clca3a1, Clca3a2), as well as GABA receptors (Gabra4, Gabrb1), cyclic nucleotide phosphodiesterases (Pde8b, Pde11a), and the neuropeptide Npy. Oct4 overexpression also selectively downregulates multiple inflammation-associated genes (B2m, H2-D1), small GTPases (Gbp2, Igtp, Irgm2, Rab13, Rgs20, Tgtp2), along with the glycolytic enzyme Gapdh and many ribosomal subunits (Fig. 4h, Table ST3). This comprises a functionally diverse range of genes that are often not prominently expressed in retinal progenitors, and whose role in controlling neurogenic competence remains to be investigated.

Oct4 also differentially regulates multiple transcription factors in both control and Rbpj-deficient Müller glia. As expected, Pou5f1 mRNA is strongly upregulated (Fig. 4e,h), with increased chromatin accessibility also being observed at the endogenous Pou5f1 promoter, indicating that Oct4 regulates its self-expression (Fig. 4i, Table ST4). Similarly, Oct4 overexpression strongly upregulates the transcription factor Rfx4 and accessibility at its endogenous promoter (Fig. 4e,h,i, Table ST4). Previous studies showed that overexpression of Rfx4 is sufficient to induce neurogenesis when overexpressed in human embryonic stem cells (Choi et al., 2024; Joung et al., 2024). However, Rfx4 expression is not detectably expressed in developing mouse retinas (Clark et al., 2019), indicating that Oct4-induced neurogenesis may involve mechanisms beyond simply inducing dedifferentiation to a retinal progenitor-like state.

Several other transcription factors are only differentially regulated following Oct4 overexpression in Rbpj-deficient Müller glia. One notable upregulated transcription factor is Klf4, which cooperates with Oct4 to induce pluripotency (Takahashi and Yamanaka, 2006). In contrast, Stat1, which inhibits neurogenesis induced by Ascl1 overexpression in Müller glia (Jorstad et al., 2020), is selectively downregulated in Rbpj-deficient Müller glia (Table ST3).

Analysis of changes in chromatin accessibility induced by Oct4 reveals that the consensus motif for Oct4 and its paralogue Pit1 (Pou1f1) show the strongest increase in accessibility, followed by Rfx family motifs (Fig. 4j,Table ST5). Both control and Rbpj-deficient Müller glia show increased accessibility associated with Sox2, which directly interacts with Oct4 and cooperates to induce pluripotency (Rizzino, 2013). While Oct4 does not significantly upregulate the pluripotency factor Klf4 in control Müller glia, it robustly induces Klf4 expression Rbpj-deficient Müller glia, and likewise selectively induces accessibility at Klf4 consensus motif sites (Fig. 4j). In addition, Rbpj-deficient Müller glia also show smaller but still significant increases in accessibility at target sites for the proneural factor Neurog2 and the neurogenic factor Insm1 (Lyu et al., 2021), identifying additional mechanisms by which Oct4 might synergistically enhance neurogenesis (Fig. 4j, Table ST5).

Discussion

In this study, we demonstrate that AAV-mediated overexpression of Oct4 is sufficient to induce neurogenic competence in Müller glia in adult mice, and to synergistically enhance neurogenesis induced by genetic disruption of Notch signaling. While Oct4 overexpression induces very low levels of proliferation, Oct4-induced neurogenesis appears to mostly result from direct glia-to-bipolar cell transdifferentiation by simultaneously upregulating expression of neurogenic bHLH factors and the bipolar cell promoting factor Otx2. In addition, Oct4 upregulates expression of the neurogenic transcription factor Rfx4, which is not expressed during retinal development. Oct4 exerts reprogramming effects by broadly increasing chromatin accessibility at predicted regulatory sites associated with these genes.

While we did not detect endogenous Oct4 expression in either injured or neurogenic mouse Müller glia, the fact that these cells express robust levels of Sox2, Klf4, and Myc implies that they may already be prime to reprogramming induced by Oct4. Our findings differ from two previous reports that initially motivated us to carry out this study. A previous study in zebrafish showed weak levels of the Oct4 homolog pou5f3 in resting glia, with these levels increasing dramatically following injury (Sharma et al., 2019). It remains unclear why we were unable to detect more than trace levels of pou5f3 expression in zebrafish Müller glia in any of the many bulk and single-cell RNA-Seq samples we analyzed (Hoang et al., 2020). We also did not detect the previously reported low levels of injury-induced Oct4 expression in mouse Müller glia (Reyes-Aguirre and Lamas, 2016). Similar discrepancies in observed cellular patterns of Oct4 expression have been observed in many other tissues (Lengner et al., 2008), and further work will be needed to clarify these discrepancies.

This study represents a clear instance in which viral-mediated expression coupled with genetic cell lineage analysis was effectively used to induce the conversion of adult mammalian Müller glia into retinal neurons in situ. A similar approach has been also successfully used to generate bipolar neurons from both immature (Pollak et al., 2013) and very recently also in mature, Müller glia by overexpression of the neurogenic bHLH factors Ascl1 and Atoh1 (Pavlou et al., 2024). Other studies reported efficient conversion of Müller glia to photoreceptors using a complex cocktail of overexpressed factors, including Ctnnb1, Otx2, Crx, and Nrl (Yao et al., 2018), and efficient conversion of Müller glia to retinal ganglion cells following overexpression of Atoh7 and Pou4f2 (Xiao et al., 2021). However, these studies both used GFAP minipromoters to both drive expression of these constructs and to infer cell lineage. Multiple studies have shown that GFAP minipromoter specificity is heavily influenced by insert sequences present in AAVs, which can result in both silencing of expression in glia and ectopic activation of expression in neurons (Le et al., 2022; L.-L. Wang et al., 2021). As a result, it cannot be concluded that these represent bona fide glia-to-neuron conversion in vivo. Similar claims have been made about knockdown of the RNA binding protein Ptbp1, which was variously claimed to directly convert Müller glia to either retinal ganglion cells or photoreceptors, but has been debunked using genetic cell lineage analysis (Hoang et al., 2022; Xie et al., 2022). In this study, the combination of prospective cell lineage analysis using the GlastCreER;Sun1-GFP transgene before infection with GFAP-Oct4-mCherry vector, in combination with scRNA-Seq analysis of infected cells, rules out any potentially confounding factors resulting from the GFAP minipromoter.

Our finding that Oct4 can synergistically enhance neurogenesis induced by genetic disruption of Notch signaling raises the question of whether this might be more broadly applicable to other models of induced neurogenesis in mammalian Müller glia. The mechanism by which Oct4 induces glial-derived neurogenesis appears complex, and to differ fundamentally from that seen following either overexpression of Ascl1 or loss of function of Nfia/b/x and/or Rbpj (Hoang et al., 2020; Jorstad et al., 2017; Le et al., 2024). In these cases, Müller glia broadly upregulate proneural genes and/or downregulate Notch signaling. Oct4 instead induces expression of the neurogenic transcription factor Rfx4, which is not expressed in developing retina. While this may represent a parallel pathway to neurogenic competence that in part accounts for synergistic induction of neurogenesis seen in Rbpj-deficient Müller glia, confirmation of the neurogenic activity of Rfx4 in Müller glia requires direct experimental confirmation.

Other mechanisms enhancing Oct4-dependent neurogenesis in the Rbpj-deficient Müller glia may include enhanced induction of Klf4 expression and increased accessibility at target sites for neurogenic transcription factors such as Ascl1. While Ascl1 overexpression induces Müller glia-derived neurogenesis in adult mice, this occurs only following treatment with HDAC inhibitors (Jorstad et al., 2017), even though Ascl1 is itself a pioneer factor (Păun et al., 2023). Oct4 overexpression broadly increases the accessibility of putative cis-regulatory sequences associated with both Neurog2 and Insm1, identifying another potential mechanism of synergy.

Transient overexpression of Oct4 could potentially further enhance the neurogenic function of Ascl1 and enhance glial-derived neurogenesis more broadly. This may also be the case for pathways reported to induce glial reprogramming in other CNS regions, such as Sox2 overexpression (Niu et al., 2015). Since Oct4 overexpression only weakly and transiently induces proliferation in adult Müller glia, this reduces the potential oncogenic risk, and may ultimately prove to be relevant for the design of cell-based regenerative therapies for treating retinal dystrophies.

Methods

Mice

Mice were raised and housed in a climate-controlled pathogen-free facility on a 14/10 h light/dark cycle. Mice used in this study were GlastCreER;Sun1-GFP, which were generated by crossing the GlastCreER and Sun1-GFP lines developed by Dr. Jeremy Nathans at Johns Hopkins (de Melo et al., 2012; Mo et al., 2015), and were obtained from his group.

GlastCreER;Rbpjlox/lox;Sun1-GFP mice were generated by crossing GlastCreER;Sun1GFP with conditional Rbpjlox/lox mice (JAX #034200).

GlastCreER;Notch1lox/lox;Notch2lox/lox;Sun1-GFP mice were obtained by crossing GlastCreER;Notch1lox/lox;Sun1-GFP and GlastCreER;Notch2lox/lox;Sun1-GFP mice. GlastCreER;Nfia/b/xlox/lox;Sun1-GFP mice were generated by crossing Nfialox/lox(see below); Nfiblox/lox (Hsu et al., 2011), and Nfixlox/lox(Campbell et al., 2008) mice to GlastCreER;Sun1-GFP mice. Nfialox/lox mice were generated in the Roswell Park Gene Targeting and Transgenic Shared Resource using heterozygous targeted ES cells from EUCOMM project 38437 (KOMP). All mice used in this study contain both the GlastCreER and Sun1-GFP transgenes, allowing visualization of both Müller glia and Müller glia-derived neurons. Maintenance and experimental procedures performed on mice were in accordance with the protocol approved by the Institutional Animal Care and Use Committee (IACUC) at the Johns Hopkins School of Medicine.

Intraperitoneal tamoxifen (TAM) Injection

To induce Cre recombination, animals at ∼3 weeks of age were intraperitoneally injected with tamoxifen (Sigma-Aldrich, #H6278-50mg) in corn oil (Sigma-Aldrich, #C8267-500ML) at 1.5mg/dose for five consecutive days.

NMDA treatment

Adult mice were anesthetized with isoflurane inhalation. Two microliters of 100 mM NMDA in PBS were intravitreally injected using a microsyringe with a 33G blunt-ended needle.

EdU treatment

At 1-week post tamoxifen induction, mice were administered EdU (Abcam, #ab146186) via i.p. injections (150μl of 5mg/ml in PBS) and drinking water (0.3 mg/ml) for 7 consecutive days.

Cloning, production, and intravitreal injection of adeno-associated virus

The Addgene #50473 construct, which contains a GFAP promoter, was used in this study. The EGFP sequence was replaced by the mCherry sequence. The P2A ribosomal self-cleaving peptide is used to simultaneously express Oct4 5′ to the mCherry reporter as a single polypeptide, which is then cleaved to generate Oct4 and mCherry. The coding sequence of Oct4 was synthesized by GeneWiz. AAV constructs were packaged into AAV2.7m8 by Boston Children’s Hospital Viral Core. Following tamoxifen induction, one-month-old GlastCreER;Sun1-GFP mice were intravitreally injected with GFAP AAV constructs using a microsyringe with a 33G blunt-ended needle. The titres and injection volume for each construct are listed in table ST6.

Immunohistochemistry and imaging

Collection and immunohistochemical analysis of retinas were performed as described previously (Hoang et al., 2020). Mouse eye globes were fixed in 4% paraformaldehyde (ElectronMicroscopySciences, #15710) for 4 h at 4°C. Retinas were dissected in 1x PBS and incubated in 30% sucrose overnight at 4°C. Retinas were then embedded in OCT (VWR, #95057-838), cryosectioned at 16 μm thickness, and stored at −20°C. Sections were dried for 30 min in a 37°C incubator and washed 3 × 5 min with 0.1% Triton X-100 in PBS (PBST). EdU labeling was performed by using Click-iT EdU kit (ThermoFisher, #C10340, #C10636) following the manufacturer’s instructions. Sections were then incubated in blocking buffer (10% Horse Serum (ThermoFisher, #26050070), 0.4% Triton X-100 in 1x PBS) for 2 h at room temperature (RT) and then incubated with primary antibodies in the blocking buffer overnight at 4°C. Primary antibodies used are listed in table ST7.

Sections were washed 4 x 5 min with PBST to remove excess primary antibodies and were incubated in secondary antibodies in blocking buffer for 2 h at RT. Sections were then counterstained with DAPI in PBST, washed 4 x 5 min in PBST and mounted with ProLong Gold Antifade Mountant (Invitrogen, #P36935) under coverslips (VWR, #48404-453), air-dried, and stored at 4°C. Fluorescent images were captured using a Zeiss LSM 700 confocal microscope. Z-stack images were collected for all sections. Colocalization of markers scored only if observed in individual Z-stack images. Secondary antibodies used are listed in table ST7.

Cell quantification and statistical analysis

Otx2/GFP-positive and HuC/D/GFP-positive cells were counted and divided by the total number of GFP-positive cells from a single random whole section per retina. Each data point in the bar graphs was calculated from an individual retina. All cell quantification data were graphed and analyzed using GraphPad Prism 10. Two-way ANOVA was used for analysis between 3 or more samples of multiple groups. All results are presented as mean ± SD.

Retinal cell dissociation

Retinas were dissected in fresh ice-cold PBS and retinal cells were dissociated using an optimized protocol as previously described (Fadl et al., 2020). Each sample contains a minimum of 4 retinas from 4 animals of both sexes. Dissociated cells were resuspended in ice-cold HBAG Buffer containing Hibernate A (BrainBits, #HALF500), B-27 supplement (ThermoFisher, #17504044), and Glutamax (ThermoFisher, #35050061).

Single-cell RNA-sequencing (scRNA-seq) library preparation

ScRNA-seq libraries were prepared using dissociated retinal cells using the 10X Genomics Chromium Single Cell 3’ Reagents Kit v3.1 (10X Genomics, Pleasanton, CA). Libraries were constructed following the manufacturer’s instructions and were sequenced using Illumina NextSeq. Sequencing data were processed through the Cell Ranger 7.0.1 pipeline (10X Genomics) using default parameters.

Single-cell Multiome ATAC + GEX sequencing library preparation

ScATAC-seq and scRNA-seq libraries were prepared using FACS-isolated GFP-positive cells using the 10X Genomic Chromium Next GEM Single Cell Multiome ATAC + Gene Expression kit following the manufacturer’s instructions. Briefly, cells were spun down at 500xg for 5 min, resuspended in 100 μl of ice-cold 0.1x Lysis Buffer, lysed by pipette-mixing 4 times, and incubated on ice for 4 min total. Cells were washed with 0.5 ml of ice-cold Wash Buffer and spun down at 500xg for 5 min at 4°C. Nuclei pellets were resuspended in 10-15 μl Nuclei Buffer and counted using Trypan blue. Resuspended cell nuclei (10-15k) were utilized for transposition and loaded into the 10x Genomics Chromium Single Cell system. ATAC libraries were amplified with 10 PCR cycles and were sequenced on Illumina NovaSeq with ∼200 million reads per library. RNA libraries were amplified from cDNA with 14 PCR cycles and were sequenced on Illumina NovaSeq 6000

scRNA-seq data analysis

ScRNA-Seq data were pre-processed using Cellranger v7.1.0 using a custom mouse genome (version mm10) with mCherry and Sun1GFP sequences included. The cell-by-genes count matrices were further analyzed using Seurat V5.1.0 (Hao et al., 2024). Cells with RNA counts less than 1000 or greater than 25,000, and number of genes less than 500 or greater than 6000 were filtered out as low-quality cells. Cells with a mitochondrial fraction of greater than 10% were removed. For data visualization, UMAP was generated using the first 20 dimensions.

Müller glia and MGPC cell clusters were subsetted for further analysis. Differential gene expression of mCherry control and Oct4-mCherry groups was performed using the “FindAllMarkers” function.

Single-cell Multiomic analysis

Raw scRNA-seq and scATAC-seq data were processed with the Cell Ranger software version 2.0.2 specifically using cellranger-arc pipeline for formatting reads, demultiplexing samples, genomic alignment, and generating the cell-by-gene count matrix and the fragments files. Both the cell-by-gene count matrix and the fragments files were the final output from the cellranger-arc pipeline, and were used for all downstream analysis.

ScRNA-seq and scATAC-seq were analyzed using Archr version 1.0.2. EnsDb.Mmusculus.v79 and BSgenome.Mmusculus.UCSC.mm10 were used for annotations. The arrow file for each sample was created using the cell-by-gene count matrix and the fragments files output from the cellranger-arc pipeline. To filter poor quality cells, Gex_nGenes greater than 500 and less than 5000 were kept, Gex_nUMI greater than 1000 and less than 15,000 were kept. TSSEnrichment greater than 10 and nFrags greater than 5000 were kept, while other cells were deemed of poor quality and were filtered out. Gene expression values from paired scATAC-seq and scRNA-seq multi modal assay were added to the arrow files using the “addGeneExpressionMatrix” function. The LSI dimensionality reduction for both scATAC and scRNASeq was computed and combined using the “addCombinedDims” function. UMAP embedding is calculated using “addUMAP” function and clusters are identified from the reduced dimensions. Marker features were used to annotate clusters.

To identify peaks, “addReproduciblePeakSet” function was used to get insertions from coverage files, call peaks, and merge peaks and this step internally used macs software. Then the “addPeakMatrix” function was called to independently compute counts for each peak per cell.

To link peaks to genes, “getPeak2GeneLinks” function was called. The “getMarkerFeatures” function was used to get the differential gene expressions and the differential peaks expressions using the GeneExpressionMatrix and PeakMatrix calculated respectively. Peaks with motifs were annotated using the cisbp motif set using “annotatePeaks.pl” from HOMMER version 4.11 to associate the peaks with nearby genes. The full code used for the analysis is available at https://github.com/SherineAwad/OCT4scRNA-ATAC